---
title: "01-tse-construction"
title-block-banner: true
author:
  - name: Cunli Pan
  - name: Jinlong Ru
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format:
  gfm: default
reference-location: section
citation-location: document
params:
  name: "01-tse-construction"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is to construct the foundational `TreeSummarizedExperiment` (TSE) object by integrating viral abundance profiles, taxonomy, host predictions, and ecosystem classification metadata from the project's SQLite database.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "01-tse-construction")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "2af4c842-7253-4469-bc24-c5193384225e")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false

suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(data.table)
  library(DBI)
  library(RSQLite)
  library(S4Vectors)
  library(IRanges)
  library(SummarizedExperiment)
  library(TreeSummarizedExperiment)
})

# Load package utility functions
devtools::load_all(here::here())

```

## Tasks

### Task 1: Read SQLite Database
```{r}
#| label: task1-read-sqlite
#| message: true
#| warning: true

# 1.1 Connect to database
sqlite_path <- file.path(path_resource, "p0057v2.sqlite")
stopifnot(file.exists(sqlite_path))

con <- DBI::dbConnect(RSQLite::SQLite(), sqlite_path)
message("✅ Connected to SQLite: ", basename(sqlite_path))

# 1.2 List all tables
tbls <- DBI::dbListTables(con)

# 1.3 Read core tables
sample_meta_raw <- DBI::dbReadTable(con, "sample_metadata")
contig_anno_raw <- DBI::dbReadTable(con, "contig_annotation")

# 1.4 Read abundance tables
abundance_map <- c(
  counts           = "abundance_count",
  tpm              = "abundance_tpm",
  rpkm             = "abundance_rpkm",
  reads_per_base   = "abundance_reads_per_base",
  trimmed_mean     = "abundance_trimmed_mean",
  covered_fraction = "abundance_covered_fraction"
)

assay_tbls <- abundance_map[abundance_map %in% tbls]
if (!length(assay_tbls)) {
  DBI::dbDisconnect(con)
  stop("❌ No abundance tables found in SQLite")
}

assay_dfs <- lapply(assay_tbls, function(tbl_name) {
  DBI::dbReadTable(con, tbl_name)
})
names(assay_dfs) <- names(assay_tbls)

# 1.5 Read AMG tables
amg_dramv_raw <- DBI::dbReadTable(con, "amg_dramv")
amg_vibrant_raw <- DBI::dbReadTable(con, "amg_vibrant")

# 1.6 Read Host prediction tables
host_genus_raw <- DBI::dbReadTable(con, "host_genus")
host_genome_raw <- DBI::dbReadTable(con, "host_genome")

# 1.7 Read Protein annotation tables
anno_prot_dramv <- DBI::dbReadTable(con, "anno_prot_dramv")
anno_prot_eggnog <- DBI::dbReadTable(con, "anno_prot_eggnog")
anno_prot_phrog <- DBI::dbReadTable(con, "anno_prot_phrog")

# 1.8 Read ARG tables
anno_arg_abricate <- DBI::dbReadTable(con, "anno_arg_abricate")
anno_arg_deeparg <- DBI::dbReadTable(con, "anno_arg_deeparg")

# 1.9 Read IMG/VR ecosystem source (external file)
imgvr_path <- file.path(path_resource, "imgvr_source.tsv")
if (file.exists(imgvr_path)) {
  imgvr_raw <- read_tsv(imgvr_path, show_col_types = FALSE)
} else {
  warning("⚠️  imgvr_source.tsv not found")
  imgvr_raw <- NULL
}

# 1.10 Close connection
DBI::dbDisconnect(con)

# Summary
message("✅ Data loaded: ",
        nrow(sample_meta_raw), " samples, ",
        nrow(contig_anno_raw), " contigs, ",
        length(assay_tbls), " abundance tables, ",
        nrow(amg_dramv_raw) + nrow(amg_vibrant_raw), " AMG entries")
```


### Task 2: Build Assay Matrices
Convert abundance data frames to matrices (vOTU × sample).
```{r}
#| label: task2-build-assays
#| message: true
#| warning: true

# 2.1 Convert abundance data frames to matrices
# Using safe_to_matrix() from R/utils.R
assay_mats <- lapply(assay_dfs, safe_to_matrix)

# 2.2 Determine common samples and vOTUs
all_samples <- lapply(assay_mats, colnames)
common_samples <- Reduce(base::intersect, all_samples)

all_features <- lapply(assay_mats, rownames)
common_features <- Reduce(base::intersect, c(list(unique(contig_anno_raw$vOTU_id)), all_features))

stopifnot(length(common_samples) > 0, length(common_features) > 0)

# 2.3 Subset and align assays
assay_mats <- lapply(assay_mats, function(m) {
  m[common_features, common_samples, drop = FALSE]
})

# Verify all assays have same dimensions
dims <- sapply(assay_mats, dim)
stopifnot(length(unique(dims[1,])) == 1, length(unique(dims[2,])) == 1)

message("✅ Assays aligned: ", nrow(assay_mats[[1]]), " vOTUs x ",
        ncol(assay_mats[[1]]), " samples")

```



### Task 3: Build colData (Sample Metadata)

```{r}
#| label: task3-build-coldata
#| message: true
#| warning: true

# 3.1 Build colData with sample_group mapping
colData_df <- sample_meta_raw %>%
  filter(sample_id %in% common_samples) %>%
  arrange(match(sample_id, common_samples)) %>%
  mutate(
    sample_group = case_when(
      sample_source == "Baltic Sea" ~ "BS",
      sample_source == "507B" ~ "SA",
      sample_source == "1327B" ~ "IA",
      sample_source == "TASF" ~ "DA",
      TRUE ~ as.character(sample_source)
    ),
    sample_group = factor(sample_group, levels = c("BS", "SA", "IA", "DA"))
  ) %>%
  column_to_rownames("sample_id")

# 3.2 Verify alignment
stopifnot(identical(rownames(colData_df), colnames(assay_mats[[1]])))

message("✅ Sample metadata built: ", nrow(colData_df), " samples (",
        paste(levels(colData_df$sample_group), collapse = ", "), ")")
```



### Task 4: Build feature metadata (rowData)
Complete vOTU-level metadata including taxonomy, quality metrics, ecosystem classification, host predictions, and contig linkage.
####  4.1 Base taxonomy and quality metrics
```{r}
#| label: task4.1-base-metadata
#| message: true
#| warning: true


# 4.1.1 Select one representative contig for each vOTU
#
# IMPORTANT:
# The primary catalogue is defined using the length of one representative
# contig per vOTU. The representative must therefore be selected before
# applying the >=5 kb threshold.
#
# Representative-selection rule used here:
#   1. longest contig;
#   2. highest CheckV completeness if lengths are tied;
#   3. contig_id as a deterministic final tie-breaker.
#
# Do not prioritize taxonomy before length. Doing so would select a 3,905-bp
# contig for vOTU0001 instead of the verified 12,805-bp representative, and a
# 4,197-bp contig for vOTU0005 instead of the verified 5,132-bp representative.

rep_contigs <- contig_anno_raw %>%
  mutate(
    representative_length_bp =
      suppressWarnings(as.numeric(contig_length)),

    checkv_completeness_num =
      suppressWarnings(as.numeric(checkv_completeness)),

    checkv_quality_score_num =
      suppressWarnings(as.numeric(checkv_quality_score))
  ) %>%
  filter(vOTU_id %in% common_features) %>%
  group_by(vOTU_id) %>%
  arrange(
    desc(representative_length_bp),
    desc(checkv_completeness_num),
    contig_id,
    .by_group = TRUE
  ) %>%
  dplyr::slice_head(n = 1) %>%
  ungroup()

# Each vOTU must now have exactly one representative contig.
stopifnot(
  nrow(rep_contigs) == length(common_features),
  !anyDuplicated(rep_contigs$vOTU_id)
)

message(
  "✅ Representative contigs selected: ",
  nrow(rep_contigs),
  " vOTUs"
)


# 4.1.2 Calculate vOTU-level contig statistics
#
# These statistics describe all contigs assigned to a vOTU.
# total_length_bp is retained for descriptive purposes only.
# It must NOT be used to define the >=5 kb primary catalogue.

votu_stats <- contig_anno_raw %>%
  filter(vOTU_id %in% common_features) %>%
  group_by(vOTU_id) %>%
  summarise(
    n_contigs = n_distinct(contig_id, na.rm = TRUE),

    total_length_bp = sum(
      suppressWarnings(as.numeric(contig_length)),
      na.rm = TRUE
    ),

    .groups = "drop"
  )


# 4.1.3 Convert the original CheckV quality score to the CheckV category
#
# The SQLite table contains checkv_quality_score:
#   1 = Complete
#   2 = High-quality
#   3 = Medium-quality
#   4 = Low-quality
#   5 = Not-determined
#
# Do not reconstruct these categories from completeness alone. CheckV quality
# categories are not simple completeness bins.

rep_contigs <- rep_contigs %>%
  mutate(
    checkv_quality = case_when(
      checkv_quality_score_num == 1 ~ "Complete",
      checkv_quality_score_num == 2 ~ "High-quality",
      checkv_quality_score_num == 3 ~ "Medium-quality",
      checkv_quality_score_num == 4 ~ "Low-quality",
      checkv_quality_score_num == 5 ~ "Not-determined",
      TRUE ~ "Not-determined"
    ),

    checkv_quality = factor(
      checkv_quality,
      levels = c(
        "Complete",
        "High-quality",
        "Medium-quality",
        "Low-quality",
        "Not-determined"
      )
    ),

    checkv_medium_or_better =
      as.character(checkv_quality) %in%
      c("Complete", "High-quality", "Medium-quality"),

    pass_length_3kb =
      !is.na(representative_length_bp) &
      representative_length_bp >= 3000,

    pass_length_5kb =
      !is.na(representative_length_bp) &
      representative_length_bp >= 5000,

    pass_length_10kb =
      !is.na(representative_length_bp) &
      representative_length_bp >= 10000
  )


# 4.1.4 Build the vOTU-level rowData table
#
# Taxonomy, CheckV information and predicted replication strategy all come
# from the selected representative contig.

rowData_base <- rep_contigs %>%
  transmute(
    vOTU_id,

    # Representative-contig identity and length
    representative_contig_id = contig_id,
    representative_length_bp,

    # Catalogue-membership flags
    pass_length_3kb,
    pass_length_5kb,
    pass_length_10kb,

    # CheckV information
    checkv_quality_score = checkv_quality_score_num,
    checkv_quality,

    # Compatibility alias:
    # keep quality_tier temporarily because older downstream scripts may refer
    # to this name. It now contains the correct CheckV category.
    quality_tier = checkv_quality,

    checkv_medium_or_better,
    checkv_provirus,
    checkv_completeness =
      suppressWarnings(as.numeric(checkv_completeness)),
    checkv_contamination =
      suppressWarnings(as.numeric(checkv_contamination)),
    checkv_completeness_method,

    # Viral taxonomy
    realm,
    kingdom,
    phylum,
    class,
    order,
    family,
    genus,
    species,

    # geNomad and other viral evidence
    classify_type,
    genomad_virus_score =
      suppressWarnings(as.numeric(genomad_virus_score)),
    genomad_n_hallmarks =
      suppressWarnings(as.numeric(genomad_n_hallmarks)),
    virsorter2_score =
      suppressWarnings(as.numeric(virsorter2_score)),
    vibrant_type,

    # Predicted replication strategy
    lifestyle,
    bacphlip_lifestyle,

    # vContact3 information
    vc_id,
    vc_genus,
    vc_classified,
    vc_with_ref,
    vc_novel_genus
  ) %>%
  left_join(
    votu_stats,
    by = "vOTU_id"
  ) %>%
  arrange(
    match(vOTU_id, rownames(assay_mats[[1]]))
  ) %>%
  column_to_rownames("vOTU_id")


# 4.1.5 Validate the representative catalogue
#
# Expected catalogue sizes for this project:
#   >=3 kb: 2,488 vOTUs
#   >=5 kb:   962 vOTUs
#   >=10 kb:  273 vOTUs
#   CheckV medium-quality or better: 55 vOTUs

stopifnot(
  identical(
    rownames(rowData_base),
    rownames(assay_mats[[1]])
  ),

  nrow(rowData_base) == 2488,

  sum(
    rowData_base$pass_length_5kb,
    na.rm = TRUE
  ) == 962,

  sum(
    rowData_base$pass_length_10kb,
    na.rm = TRUE
  ) == 273,

  sum(
    rowData_base$checkv_medium_or_better,
    na.rm = TRUE
  ) == 55
)


# 4.1.6 Report CheckV quality-category counts

quality_summary <- rowData_base %>%
  as.data.frame() %>%
  dplyr::count(
    checkv_quality,
    name = "n_vOTUs",
    .drop = FALSE
  )

print(quality_summary)

message(
  "✅ 4.1 Base metadata completed: ",
  nrow(rowData_base),
  " discovery-catalogue vOTUs"
)

message(
  "   >=5 kb primary catalogue: ",
  sum(rowData_base$pass_length_5kb),
  " vOTUs"
)

message(
  "   >=10 kb catalogue: ",
  sum(rowData_base$pass_length_10kb),
  " vOTUs"
)

message(
  "   CheckV medium-quality or better: ",
  sum(rowData_base$checkv_medium_or_better),
  " vOTUs"
)

```



#### 4.2 Ecosystem classification
```{r}
 #| label: task4.2-ecosystem
#| message: true
#| warning: true

if (!is.null(imgvr_raw)) {

  # 4.2.1 Using extract_level5_position5() from R/utils.R

  # 4.2.2 Extract Level 5 from imgvr
  imgvr_eco <- imgvr_raw %>%
    select(contig_id, eco_class = `Ecosystem classification`) %>%
    mutate(level5 = sapply(eco_class, extract_level5_position5)) %>%
    filter(!is.na(level5)) %>%
    select(contig_id, eco_level5 = level5) %>%
    distinct()

  # 4.2.3 Get contig_id for each vOTU
  votu_to_contig <- rep_contigs %>%
    filter(vOTU_id %in% rownames(rowData_base)) %>%
    select(vOTU_id, contig_id)

  # 4.2.4 Merge ecosystem info
  votu_eco <- votu_to_contig %>%
    left_join(imgvr_eco, by = "contig_id") %>%
    select(vOTU_id, eco_level5)

  # 4.2.5 Add to rowData_base
  rowData_base <- rowData_base %>%
    rownames_to_column("vOTU_id") %>%
    left_join(votu_eco, by = "vOTU_id") %>%
    arrange(match(vOTU_id, rownames(assay_mats[[1]]))) %>%
    column_to_rownames("vOTU_id")

  n_with_eco <- sum(!is.na(rowData_base$eco_level5))
  message("✅ 4.2 Ecosystem: ", n_with_eco, "/", nrow(rowData_base),
          " vOTUs (", round(n_with_eco/nrow(rowData_base)*100, 1), "%)")

} else {
  warning("⚠️  Skipping ecosystem annotation (imgvr_source.tsv not loaded)")
}

stopifnot(identical(rownames(rowData_base), rownames(assay_mats[[1]])))

```



#### 4.3 Host predictions

```{r}
#| label: task4.3-host-predictions
#| message: true
#| warning: true

# 4.3.1 Aggregate host genus predictions
host_genus_agg <- host_genus_raw %>%
  filter(vOTU_id %in% rownames(rowData_base)) %>%
  group_by(vOTU_id) %>%
  summarise(
    hostGenusset = paste(unique(Host.genus), collapse = ";"),
    hostnumbergenus = n_distinct(Host.genus),
    .groups = "drop"
  ) %>%
  mutate(
    hostSG = case_when(
      hostnumbergenus == 1 ~ "specialist",
      hostnumbergenus > 1 ~ "generalist",
      TRUE ~ NA_character_
    )
  )

# 4.3.2 Aggregate host species predictions
host_species_agg <- host_genome_raw %>%
  filter(vOTU_id %in% rownames(rowData_base)) %>%
  group_by(vOTU_id) %>%
  summarise(
    hostSpeciesset = paste(unique(Host.taxonomy), collapse = ";"),
    hostnumberspecies = n_distinct(Host.taxonomy),
    .groups = "drop"
  )

# 4.3.3 Merge host info into rowData
rowData_base <- rowData_base %>%
  rownames_to_column("vOTU_id") %>%
  left_join(host_genus_agg, by = "vOTU_id") %>%
  left_join(host_species_agg, by = "vOTU_id") %>%
  arrange(match(vOTU_id, rownames(assay_mats[[1]]))) %>%
  column_to_rownames("vOTU_id")

stopifnot(identical(rownames(rowData_base), rownames(assay_mats[[1]])))

n_with_host <- sum(!is.na(rowData_base$hostGenusset))
message("✅ 4.3 Host predictions: ", n_with_host, "/", nrow(rowData_base),
        " vOTUs (", round(n_with_host/nrow(rowData_base)*100, 1), "%)")
```
#### 4.4 Contig linkage
```{r}
#| label: task4.4-contig-linkage
#| message: true
#| warning: true

# 4.4.1 Build vOTU → contig mapping
bridge_votu_contig <- contig_anno_raw %>%
  filter(!is.na(vOTU_id), !is.na(contig_id),
         vOTU_id != "", contig_id != "",
         vOTU_id %in% rownames(rowData_base)) %>%
  distinct(vOTU_id, contig_id) %>%
  arrange(vOTU_id, contig_id)

# 4.4.2 Create CharacterList
v2c_map <- split(bridge_votu_contig$contig_id, bridge_votu_contig$vOTU_id)
v2c_map <- lapply(v2c_map, unique)

lookup <- function(id) {
  x <- v2c_map[[id]]
  if (is.null(x)) character(0) else as.character(x)
}

votu_ids_ordered <- rownames(rowData_base)
contigs_aligned <- lapply(votu_ids_ordered, lookup)
contig_ids_list <- IRanges::CharacterList(contigs_aligned)
names(contig_ids_list) <- votu_ids_ordered

# 4.4.3 Add to rowData
rowData_base$contig_ids <- contig_ids_list

stopifnot(identical(rownames(rowData_base), rownames(assay_mats[[1]])))

message("✅ 4.4 Contig linkage: ", nrow(bridge_votu_contig), " edges, ",
        sum(lengths(contig_ids_list) > 1), " multi-contig vOTUs")

# Final rowData summary
message("✅ Task 4 completed: rowData with ", ncol(rowData_base), " columns")

```


### Task 5: Task 5: Construct and validate TSE object
Prepare metadata, construct TreeSummarizedExperiment, and validate output.
#### 5.1 Prepare metadata list
```{r}
#| label: task5.1-prepare-metadata
#| message: true
#| warning: true

# Prepare complete metadata list
metadata_list <- list(
  # Construction info
  construction_info = list(
  created_date = Sys.Date(),
  created_time = Sys.time(),
  sqlite_path = normalizePath(sqlite_path),

  n_samples_raw = nrow(sample_meta_raw),
  n_samples = length(common_samples),

  n_votus_raw =
    length(unique(contig_anno_raw$vOTU_id)),

  n_votus_discovery =
    nrow(rowData_base),

  n_votus_primary_5kb =
    sum(rowData_base$pass_length_5kb),

  n_votus_10kb =
    sum(rowData_base$pass_length_10kb),

  n_votus_checkv_medium_or_better =
    sum(rowData_base$checkv_medium_or_better),

  assays_loaded =
    names(assay_mats),

  strategy_representative_contig =
    "longest representative contig > CheckV completeness > contig_id",

  primary_catalogue_definition =
    "representative contig length >= 5000 bp",

  permissive_detection_definition =
    "TPM > 0",

  higher_confidence_detection_definition =
    "covered fraction >= 0.5"
  ),

  # Protein annotations
  anno_prot_dramv = anno_prot_dramv,
  anno_prot_eggnog = anno_prot_eggnog,
  anno_prot_phrog = anno_prot_phrog,

  # AMG tables
  amg_dramv = amg_dramv_raw,
  amg_vibrant = amg_vibrant_raw,

  # ARG tables
  anno_arg_abricate = anno_arg_abricate,
  anno_arg_deeparg = anno_arg_deeparg,

  # Host prediction edge tables
  host_genus_edges = host_genus_raw,
  host_genome_edges = host_genome_raw,

  # Contig linkage info
  linkage_info = list(
    n_votu_contig_edges = nrow(bridge_votu_contig),
    multi_contig_votus = names(v2c_map)[lengths(v2c_map) > 1],
    n_multi_contig_votus = sum(lengths(v2c_map) > 1)
  )
)

# Add complete contig annotation
metadata_list$contig_annotation <- contig_anno_raw %>%
  filter(vOTU_id %in% rownames(rowData_base))

# Add complete imgvr source if available
if (!is.null(imgvr_raw)) {
  metadata_list$imgvr_source <- imgvr_raw
}

message("✅ 5.1 Metadata prepared: ", length(metadata_list), " components")
```



#### 5.2 Construct TSE
```{r}
#| label: task5.2-construct-tse
#| message: true
#| warning: true
#| 
stopifnot(
  identical(
    rownames(assay_mats[[1]]),
    rownames(rowData_base)
  ),

  identical(
    colnames(assay_mats[[1]]),
    rownames(colData_df)
  )
)

message("✅ 5.2 Input dimensions verified")


# 5.2.2 Construct the complete discovery-catalogue TSE
#
# This object contains all 2,488 vOTUs. It is retained for sensitivity
# analyses involving the >=3 kb catalogue and genome-recovery effects.

tse_discovery <- TreeSummarizedExperiment(
  assays = S4Vectors::SimpleList(assay_mats),
  colData = S4Vectors::DataFrame(colData_df),
  rowData = S4Vectors::DataFrame(rowData_base),
  metadata = metadata_list
)

stopifnot(
  nrow(tse_discovery) == 2488,
  ncol(tse_discovery) == 4
)

metadata(tse_discovery)$catalogue_info <- list(
  catalogue_name = "Discovery catalogue",
  catalogue_size = 2488,
  representative_length_threshold_bp = 3000,
  tpm_normalisation =
    "Original TPM normalisation across the 2,488-vOTU catalogue"
)

message(
  "✅ Discovery TSE constructed: ",
  nrow(tse_discovery),
  " vOTUs × ",
  ncol(tse_discovery),
  " samples"
)


# 5.2.3 Subset the >=5 kb primary catalogue

primary_ids <- rownames(tse_discovery)[
  rowData(tse_discovery)$pass_length_5kb
]

stopifnot(
  length(primary_ids) == 962,
  !anyDuplicated(primary_ids)
)

tse_primary <- tse_discovery[
  primary_ids,
  ,
  drop = FALSE
]

stopifnot(
  nrow(tse_primary) == 962,
  ncol(tse_primary) == 4
)


# 5.2.4 Preserve the source TPM values
#
# The SQLite TPM matrix was normalised across all 2,488 discovery vOTUs.
# After subsetting to 962 vOTUs, preserve those source values under an
# unambiguous assay name.

assay(
  tse_primary,
  "tpm_source_2488"
) <- assay(
  tse_primary,
  "tpm"
)


# 5.2.5 Renormalise TPM within the 962-vOTU primary catalogue

tpm_primary_source <- assay(
  tse_primary,
  "tpm_source_2488"
)

tpm_primary_totals <- colSums(
  tpm_primary_source,
  na.rm = TRUE
)

stopifnot(
  all(is.finite(tpm_primary_totals)),
  all(tpm_primary_totals > 0)
)

tpm_primary_renormalised <- sweep(
  tpm_primary_source,
  2,
  tpm_primary_totals,
  "/"
) * 1e6

assay(
  tse_primary,
  "tpm"
) <- tpm_primary_renormalised


# 5.2.6 Record the primary-catalogue definition

metadata(tse_primary)$catalogue_info <- list(
  catalogue_name = "Primary catalogue",
  catalogue_size = 962,
  representative_length_threshold_bp = 5000,

  representative_selection =
    "longest contig > CheckV completeness > contig_id",

  tpm_assay =
    "Renormalised within the retained 962-vOTU catalogue to 1e6 per sample",

  tpm_source_2488_assay =
    "Original source TPM values inherited from the 2,488-vOTU catalogue",

  counts_assay =
    "Original mapped read counts; not renormalised",

  covered_fraction_assay =
    "Original covered fraction; not renormalised"
)


# 5.2.7 Validate TPM normalisation

stopifnot(
  all(
    abs(
      colSums(
        assay(tse_primary, "tpm"),
        na.rm = TRUE
      ) - 1e6
    ) < 1
  )
)

message(
  "✅ Primary TSE constructed: ",
  nrow(tse_primary),
  " vOTUs × ",
  ncol(tse_primary),
  " samples"
)

message(
  "   Discovery assays: ",
  paste(
    assayNames(tse_discovery),
    collapse = ", "
  )
)

message(
  "   Primary assays: ",
  paste(
    assayNames(tse_primary),
    collapse = ", "
  )
)
```



#### 5.3 Save and validate
```{r}
#| label: task5.3-save-validate
#| message: true
#| warning: true

# 5.3.1 Save TSE
# 5.3.1 Define explicit output paths
#
# Do not use an ambiguous file name such as tse.rds. Downstream analyses must
# state whether they use the complete discovery catalogue or the primary
# >=5 kb catalogue.

discovery_output_path <- path_target(
  "tse_discovery_2488.rds"
)

primary_output_path <- path_target(
  "tse_primary_962.rds"
)


# 5.3.2 Save both TSE objects

saveRDS(
  tse_discovery,
  discovery_output_path
)

saveRDS(
  tse_primary,
  primary_output_path
)


# 5.3.3 Export a compact catalogue manifest
#
# This CSV contains one row per discovery-catalogue vOTU and records which
# filtering scenarios it belongs to.

catalogue_manifest <- rowData(tse_discovery) %>%
  as.data.frame() %>%
  rownames_to_column("vOTU_id") %>%
  dplyr::select(
    vOTU_id,
    representative_contig_id,
    representative_length_bp,

    checkv_completeness,
    checkv_quality_score,
    checkv_quality,
    checkv_medium_or_better,

    pass_length_3kb,
    pass_length_5kb,
    pass_length_10kb,

    realm,
    kingdom,
    phylum,
    class,
    order,
    family,
    genus,
    species,

    lifestyle,
    bacphlip_lifestyle,

    n_contigs,
    total_length_bp
  )

write_csv(
  catalogue_manifest,
  path_target("votu_catalogue_manifest_2488.csv")
)


# 5.3.4 Export the primary-catalogue manifest

primary_manifest <- catalogue_manifest %>%
  filter(pass_length_5kb)

stopifnot(
  nrow(primary_manifest) == 962
)

write_csv(
  primary_manifest,
  path_target("votu_primary_manifest_962.csv")
)


# 5.3.5 Save session information

session_info_path <- path_target(
  "session_info.txt"
)

writeLines(
  capture.output(sessionInfo()),
  session_info_path
)


# 5.3.6 Reload both files and validate the saved objects

tse_discovery_reload <- readRDS(
  discovery_output_path
)

tse_primary_reload <- readRDS(
  primary_output_path
)

validation_passed <- all(
  identical(
    dim(tse_discovery_reload),
    c(2488L, 4L)
  ),

  identical(
    dim(tse_primary_reload),
    c(962L, 4L)
  ),

  "amg_dramv" %in%
    names(metadata(tse_discovery_reload)),

  "host_genome_edges" %in%
    names(metadata(tse_discovery_reload)),

  "representative_contig_id" %in%
    colnames(rowData(tse_discovery_reload)),

  "representative_length_bp" %in%
    colnames(rowData(tse_discovery_reload)),

  "pass_length_5kb" %in%
    colnames(rowData(tse_discovery_reload)),

  "checkv_quality" %in%
    colnames(rowData(tse_discovery_reload)),

  "sample_group" %in%
    colnames(colData(tse_discovery_reload)),

  "tpm_source_2488" %in%
    assayNames(tse_primary_reload),

  all(
    abs(
      colSums(
        assay(tse_primary_reload, "tpm"),
        na.rm = TRUE
      ) - 1e6
    ) < 1
  )
)

if (!validation_passed) {
  stop("❌ Validation failed")
}

message(
  "✅ Validation passed: both TSE objects are ready"
)


# 5.3.7 Display final summary

message("\n=== Final TSE Summary ===")

message(
  "Discovery catalogue: ",
  nrow(tse_discovery_reload),
  " vOTUs × ",
  ncol(tse_discovery_reload),
  " samples"
)

message(
  "Primary catalogue: ",
  nrow(tse_primary_reload),
  " vOTUs × ",
  ncol(tse_primary_reload),
  " samples"
)

message(
  "Sample groups: ",
  paste(
    levels(colData(tse_primary_reload)$sample_group),
    collapse = ", "
  )
)

message(
  "Discovery assays: ",
  paste(
    assayNames(tse_discovery_reload),
    collapse = ", "
  )
)

message(
  "Primary assays: ",
  paste(
    assayNames(tse_primary_reload),
    collapse = ", "
  )
)

message(
  "Primary TPM column sums: ",
  paste(
    round(
      colSums(
        assay(tse_primary_reload, "tpm")
      ),
      2
    ),
    collapse = ", "
  )
)

message("\n✅ Files saved:")

message(
  "   - ",
  basename(discovery_output_path)
)

message(
  "   - ",
  basename(primary_output_path)
)

message(
  "   - votu_catalogue_manifest_2488.csv"
)

message(
  "   - votu_primary_manifest_962.csv"
)

message(
  "   - ",
  basename(session_info_path)
)
```

