---
title: "04-fig4-heatmap"
title-block-banner: true
author:
  - name: Cunli Pan
  - name: Jinlong Ru
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format:
  gfm: default
reference-location: section
citation-location: document
params:
  name: "04-fig4-heatmap"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is to construct and visualize a heatmap of viral functional potential based on KEGG orthology (KO) and pathway annotations to infer metabolic capabilities of the viral community.

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "04-fig4-heatmap")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "008e69a8-ea7d-4750-9b7b-10b076a2b11a")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(TreeSummarizedExperiment)
  library(ComplexHeatmap)
  library(circlize)
  library(KEGGREST)
  library(openxlsx)
  library(progress)
  library(viridis)
   library(RColorBrewer)
  library(pheatmap)

})

# Load package utility functions
devtools::load_all(here::here())

```

## Tasks

### Task 1: Data-extraction

```{r}
#| label: data-extraction
# ============================================================================
# STEP 1: 从TSE加载数据
# ============================================================================
cat("Part 1: Loading TSE object...\n\n")

tse_path <- here::here(
  "data",
  "01-tse-construction",
  "tse_primary_962.rds"
)

stopifnot(file.exists(tse_path))

tse <- readRDS(tse_path)

stopifnot(
  nrow(tse) == 962,
  ncol(tse) == 4
)

colData(tse)$sample_group <- factor(
  colData(tse)$sample_group,
  levels = c("BS", "SA", "IA", "DA")
)

# TSE metadata does not automatically follow row subsetting,
# so filter the candidate tables explicitly.
amg_dramv <- metadata(tse)$amg_dramv %>%
  dplyr::filter(vOTU_id %in% rownames(tse))

amg_vibrant <- metadata(tse)$amg_vibrant %>%
  dplyr::filter(vOTU_id %in% rownames(tse))

n_unique_candidates <- dplyr::bind_rows(
  amg_dramv %>% dplyr::select(vOTU_id, protein_id),
  amg_vibrant %>% dplyr::select(vOTU_id, protein_id)
) %>%
  dplyr::distinct() %>%
  nrow()

stopifnot(
  nrow(amg_dramv) == 72,
  nrow(amg_vibrant) == 124,
  n_unique_candidates == 167
)

message("TSE: ", nrow(tse), " vOTUs × ", ncol(tse), " samples")
message("DRAM-v candidate rows: ", nrow(amg_dramv))
message("VIBRANT candidate rows: ", nrow(amg_vibrant))
message("Unique T1 candidate ORFs: ", n_unique_candidates)
```

### Task 2: Annotate KEGG

```{r}
#| label: annotate-kegg
#| message: true
#| warning: false

message("=== Annotate AMG Genes with KEGG ===")

# Extract unique dbids
dramv_ids <- amg_dramv %>%
  dplyr::select(dbid) %>%
  dplyr::filter(!is.na(dbid)) %>%
  dplyr::distinct() %>%
  mutate(source = "DRAM-v")

vibrant_ids <- amg_vibrant %>%
  dplyr::select(dbid) %>%
  dplyr::filter(!is.na(dbid)) %>%
  dplyr::distinct() %>%
  mutate(source = "VIBRANT")

all_ids <- bind_rows(dramv_ids, vibrant_ids) %>%
  dplyr::distinct(dbid)

message("Total unique dbids: ", nrow(all_ids))

# KEGG annotation function
# Using get_kegg_info() from R/utils.R

# Annotate all dbids
annotations <- purrr::map_dfr(all_ids$dbid, get_kegg_info)

message("Annotation completed: ", nrow(annotations), " dbids")
message("  With pathway: ", sum(!is.na(annotations$Pathway)))

# Merge annotations
dramv_annotated <- dplyr::left_join(amg_dramv, annotations, by = "dbid")
vibrant_annotated <- dplyr::left_join(amg_vibrant, annotations, by = "dbid")

# Export
write.xlsx(annotations, path_target("all_kegg_annotations.xlsx"))
write.xlsx(dramv_annotated, path_target("dramv_with_annotations.xlsx"))
write.xlsx(vibrant_annotated, path_target("vibrant_with_annotations.xlsx"))

message("✅ Task 2 completed")

```


### Task 3: Pathway-split-classify

```{r}
#| label: pathway-split-classify
#| message: true
#| warning: false

message("=== Pathway Splitting and Classification ===")

# Read annotated AMG tables
dramv_anno <- read.xlsx(path_target("dramv_with_annotations.xlsx"))
vibrant_anno <- read.xlsx(path_target("vibrant_with_annotations.xlsx"))

# Combine and add source label
combined_raw <- bind_rows(
  dramv_anno %>% mutate(Source = "DRAM-v"),
  vibrant_anno %>% mutate(Source = "VIBRANT")
)

original_row_count <- nrow(combined_raw)
message("Original rows: ", original_row_count)

# Split pathways (core step)
long_pathway <- combined_raw %>%
  mutate(
    has_pathway = !is.na(Pathway) & Pathway != "",
    row_id = row_number()
  ) %>%
  mutate(
    Pathway_ID = ifelse(has_pathway, Pathway, NA_character_)
  ) %>%
  separate_rows(Pathway_ID, sep = ";\\s*", convert = FALSE) %>%
  mutate(
    Pathway_ID = str_trim(Pathway_ID),
    Pathway_ID = ifelse(Pathway_ID == "", NA_character_, Pathway_ID),
    is_expanded = ifelse(has_pathway & !is.na(Pathway_ID), TRUE, FALSE)
  )

expanded_row_count <- nrow(long_pathway)
message("Expanded rows: ", expanded_row_count, " (+", expanded_row_count - original_row_count, ")")

# Get unique pathway IDs
pathway_ids <- unique(na.omit(long_pathway$Pathway_ID))
total_ids <- length(pathway_ids)
message("Unique pathways to query: ", total_ids)

# KEGG pathway query function
# Using get_pathway_info() from R/utils.R

# Batch query KEGG API
message("Starting batch KEGG query...")
pathway_annotations <- tibble()

pb <- progress_bar$new(
  format = "[:bar] :percent ETA: :eta",
  total = total_ids,
  clear = FALSE
)

for (i in seq_along(pathway_ids)) {
  id <- pathway_ids[i]
  res <- get_pathway_info(id, i, total_ids)
  pathway_annotations <- bind_rows(pathway_annotations, res)
  pb$tick()

  if (i %% 10 == 0) Sys.sleep(0.3)
}

# Merge pathway annotations
long_pathway_annotated <- long_pathway %>%
  dplyr::left_join(pathway_annotations, by = "Pathway_ID") %>%
  mutate(
    annotation_status = case_when(
      is.na(Pathway_ID) ~ "No Pathway",
      is.na(Pathway_Name) ~ "Missing Annotation",
      str_starts(Pathway_Name, "Unknown") ~ "API Failed",
      TRUE ~ "Annotated"
    )
  )

# Save results
write.xlsx(long_pathway_annotated, path_target("long_pathway_with_annotation.xlsx"))
message("Saved: long_pathway_with_annotation.xlsx")

pathway_class_ref <- pathway_annotations %>%
  dplyr::distinct(Pathway_ID, Pathway_Name, Pathway_Class, Pathway_Top_Category)
write.xlsx(pathway_class_ref, path_target("pathway_classification_reference.xlsx"))
message("Saved: pathway_classification_reference.xlsx")

# Statistics
message("\nFinal statistics:")
message("  Successful queries: ", sum(!str_starts(pathway_annotations$Pathway_Name, "Unknown")), "/", total_ids)
print(table(pathway_annotations$Pathway_Top_Category))

message("\n=== Task 3 completed ===\n")

```


### Task 4: Merge TPM and Metadata

```{r}
#| label: merge-tpm-metadata
#| message: true
#| warning: false

message("=== Merge TPM and Metadata ===")

# Extract TPM matrix from TSE
tpm_matrix <- assays(tse)$tpm
sample_meta <- colData(tse) %>% as_tibble(rownames = "sample_id")

# Convert TPM to long format
tpm_long <- tpm_matrix %>%
  as.data.frame() %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM")

# Merge sample metadata
tpm_long_meta <- tpm_long %>%
  dplyr::left_join(sample_meta, by = "sample_id") %>%
   mutate(sample_group = as.character(sample_group))

# Read pathway table
long_pathway <- read.xlsx(path_target("long_pathway_with_annotation.xlsx"))

# Merge all information
final_df <- long_pathway %>%
  dplyr::left_join(tpm_long_meta, by = "vOTU_id")

message("Final merged table: ", nrow(final_df), " rows")

# Save
write.xlsx(final_df, path_target("full_merged_pathway_table.xlsx"))
message("✅ Task 4 completed")
```


### Task 5: Deduplicate and Calculate TPM

```{r}
#| label: dedup-calculate-tpm
#| message: true
#| warning: false

message("=== Deduplicate and Calculate Pathway Total TPM ===")

# Read full merged table
full_df <- read.xlsx(path_target("full_merged_pathway_table.xlsx"))

# Deduplicate
deduplicated_df <- full_df %>%
  arrange(vOTU_id, dbid, Pathway_ID, sample_id,
          match(Source, c("DRAM-v", "VIBRANT"))) %>%
  dplyr::distinct(vOTU_id, dbid, Pathway_ID, sample_id, .keep_all = TRUE)

message("Deduplicated: ", nrow(full_df), " → ", nrow(deduplicated_df))
write.xlsx(deduplicated_df, path_target("deduplicated_pathway_table.xlsx"))

# Filter valid pathways
# Restrict pathway-level visualization to categories applicable to microbial/viral genomes
# Filter valid pathways
# Restrict pathway-level visualization to KEGG Metabolism pathways
filtered_df <- deduplicated_df %>%
  dplyr::filter(!is.na(Pathway_ID), !is.na(Pathway_Name), !is.na(sample_group)) %>%
  dplyr::filter(!is.na(TPM)) %>%
  dplyr::filter(Pathway_Top_Category == "Metabolism") %>%
  mutate(
  Pathway_Group = case_when(
    grepl("^map01[123]", Pathway_ID) ~ "Global and overview maps",
    !is.na(Pathway_Class) & Pathway_Class != "" ~
      sub("^Metabolism;\\s*", "", Pathway_Class),
    TRUE ~ "Unclassified metabolic pathway"
  )
)


message("Valid pathway rows: ", nrow(filtered_df))

# Calculate total TPM per pathway
samples <- c("BS", "SA", "IA", "DA")

heatmap_matrix <- filtered_df %>%
  group_by(Pathway_ID, Pathway_Name, Pathway_Group, sample_group) %>%
  summarise(Total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sample_group, values_from = Total_TPM, values_fill = 0)

final_df_stats <- heatmap_matrix %>%
  mutate(
    Total_TPM = rowSums(across(all_of(samples))),
    Samples_Expressed = rowSums(across(all_of(samples), ~ .x > 0))
  ) %>%
  arrange(desc(Total_TPM))

message("Heatmap matrix: ", nrow(final_df_stats), " pathways")
write.xlsx(final_df_stats, path_target("heatmap_matrix_with_pathway_group.xlsx"))
message("✅ Task 5 completed")
```




### Task 6: Plot Pathway Heatmap

```{r}
#| label: plot-pathway-heatmap
#| message: true
#| warning: false
#| fig.width: 10
#| fig.height: 12

message("=== Plot Pathway Heatmap ===")

# Shared heatmap color scale
shared_heatmap_colors <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))(100)
shared_heatmap_breaks <- seq(0, 6, length.out = 101)  # for pheatmap
shared_heatmap_values <- seq(0, 6, length.out = length(shared_heatmap_colors))  # for ComplexHeatmap

shared_heatmap_col_fun <- circlize::colorRamp2(
  shared_heatmap_values,
  shared_heatmap_colors
)
# Read data
df <- read.xlsx(path_target("heatmap_matrix_with_pathway_group.xlsx"))
samples <- c("BS", "SA", "IA", "DA")

# Extract pathway short name
df <- df %>%
  mutate(Pathway_Short = sub(";.*", "", Pathway_Name))

# Convert to log10(TPM + 1)
df_heatmap <- df %>%
  dplyr::select(Pathway_Short, all_of(samples)) %>%
  mutate(across(all_of(samples), ~log10(. + 1))) %>%
  column_to_rownames("Pathway_Short")

df_heatmap <- df_heatmap[, samples]

message("Plotting: ", nrow(df_heatmap), " pathways × ", ncol(df_heatmap), " samples")

# KEGG pathway-class annotation
pathway_class_levels <- unique(df$Pathway_Group)

pathway_class_colours <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(12, "Set3"))(
    length(pathway_class_levels)
  ),
  pathway_class_levels
)

legend_labels <- c(
  "Global and overview maps" = "Global and overview maps",
  "Energy metabolism" = "Energy metabolism",
  "Amino acid metabolism" = "Amino acid metabolism",
  "Nucleotide metabolism" = "Nucleotide metabolism",
  "Xenobiotics biodegradation and metabolism" = "Xenobiotic metabolism",
  "Glycan biosynthesis and metabolism" = "Glycan metabolism",
  "Biosynthesis of other secondary metabolites" = "Other secondary metabolites",
  "Metabolism of terpenoids and polyketides" = "Terpenoid/polyketide metabolism",
  "Metabolism of cofactors and vitamins" = "Cofactor and vitamin metabolism",
  "Carbohydrate metabolism" = "Carbohydrate metabolism",
  "Lipid metabolism" = "Lipid metabolism",
  "Metabolism of other amino acids" = "Other amino acid metabolism",
  "Unclassified metabolic pathway" = "Unclassified metabolic pathway"
)

pathway_annotation <- rowAnnotation(
  `KEGG class` = factor(
    df$Pathway_Group,
    levels = pathway_class_levels
  ),
  col = list(
    `KEGG class` = pathway_class_colours
  ),
  show_annotation_name = FALSE,
  simple_anno_size = unit(2.5, "mm"),
  annotation_legend_param = list(
    `KEGG class` = list(
      title = "KEGG pathway class",
      labels = unname(
        legend_labels[pathway_class_levels]
      ),
      direction = "vertical",
      ncol = 1,
      labels_gp = gpar(
        fontsize = 8,
        fontfamily = "Times"
      ),
      title_gp = gpar(
        fontsize = 9,
        fontfamily = "Times",
        fontface = "bold"
      )
    )
  )
)

label_colors <- "black"



# Convert to matrix
mat <- as.matrix(df_heatmap)

# Plot ComplexHeatmap
ht <- Heatmap(
  mat,
  left_annotation = pathway_annotation,
  name = "log10(TPM + 1)",
  col = shared_heatmap_col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  width = unit(10, "cm"),
  height = unit(18, "cm"),

  row_names_side = "left",
  row_names_gp = gpar(fontsize = 9, fontfamily = "Times", col = label_colors),
  row_names_max_width = unit(7, "cm"),

  show_column_names = TRUE,
  column_names_side = "bottom",
  column_names_rot = 0,
  column_names_gp = gpar(fontsize = 18, fontfamily = "Times", fontface = "plain"),

  border = TRUE,

  heatmap_legend_param = list(
    title = NULL,
    at = 0:6,
    labels = as.character(0:6),
    legend_height = unit(5, "cm"),
    direction = "vertical",
    grid_height = unit(0.3, "cm"),
    labels_gp = gpar(fontsize = 10, fontfamily = "Times")
  ),

  column_title = "Pathway Abundance Across Samples (log10(TPM + 1))",
  column_title_gp = gpar(fontsize = 12, fontfamily = "Times", fontface = "bold"),

  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(x, y, width, height,
              gp = gpar(col = "white", fill = NA, lwd = 0.3))
  }
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "bottom"
)

# Save figures
png(
  path_target("Fig4_pathway_heatmap.png"),
  width = 1000,
  height = 1000,
  res = 120
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "bottom"
)

dev.off()

pdf(
  path_target("Fig4_pathway_heatmap.pdf"),
  width = 12,
  height = 16
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "bottom"
)

dev.off()

png(
  path_target("Fig4_pathway_heatmap_highres.png"),
  width = 3600,
  height = 4800,
  res = 300
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "bottom"
)

dev.off()

# Save data
write.csv(df_heatmap %>% rownames_to_column("Pathway_Short"),
          path_target("log10_TPM_heatmap_matrix.csv"), row.names = FALSE)

message("✅ Task 6 completed")
```



### Task 7: Plot Top 20 AMG Heatmap
```{r}
#| label: plot-amg-heatmap-total
#| message: true
#| warning: false
#| fig.width: 8
#| fig.height: 10

message("=== Plot Top 20 AMG Gene Heatmap (Total TPM) ===")

# Shared heatmap color scale
shared_heatmap_colors <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))(100)
shared_heatmap_breaks <- seq(0, 6, length.out = 101)

# Read deduplicated pathway table
df <- read.xlsx(path_target("deduplicated_pathway_table.xlsx"))

# Clean data
df_clean <- df %>%
  dplyr::select(dbid, vOTU_id, protein_id, sample_group, TPM, db_desc) %>%
  dplyr::filter(!is.na(dbid), !is.na(sample_group), !is.na(TPM)) %>%
  dplyr::distinct()

message("Clean data: ", nrow(df_clean), " rows")

# Special gene name mapping
special_map <- tribble(
  ~dbid, ~gene_symbol,
  "K00525", "RNR",
  "EC:1.17.4.1", "RNR",
  "K00973", "rfbA",
  "K01710", "rfbB",
  "K00560", "thyA",
  "K01520", "dut",
  "K00558", "DNMT1",
  "K17398", "DNMT3A",
  "K00287", "folA",
  "K01494", "dcd",
  "K02935", "rplL"
)

# Extract gene labels
df_labeled <- df_clean %>%
  dplyr::left_join(special_map, by = "dbid") %>%
  mutate(
    auto_label = str_remove_all(db_desc, "\\[.*?\\]") %>%
      str_split(";", simplify = TRUE) %>%
      .[, 1] %>%
      str_trim() %>%
      str_split(",", simplify = TRUE) %>%
      .[, 1] %>%
      str_trim(),
    gene_symbol = if_else(is.na(gene_symbol), auto_label, gene_symbol)
  ) %>%
  dplyr::select(-auto_label) %>%
  # Collapse duplicate KO/EC annotations assigned to the same protein
  dplyr::distinct(vOTU_id, protein_id, gene_symbol, sample_group, TPM, .keep_all = TRUE)

# Summarize total TPM per gene
gene_summary <- df_labeled %>%
  group_by(gene_symbol, sample_group) %>%
  summarise(Total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop")

# Convert to wide format
gene_matrix <- gene_summary %>%
  pivot_wider(names_from = sample_group, values_from = Total_TPM, values_fill = 0)

# Select Top 20
samples <- c("BS", "SA", "IA", "DA")
available_samples <- intersect(samples, colnames(gene_matrix))

gene_top20_total <- gene_matrix %>%
  mutate(Sum_TPM = rowSums(across(all_of(available_samples)))) %>%
  arrange(desc(Sum_TPM)) %>%
  slice_head(n = 20) %>%
  dplyr::select(-Sum_TPM)

message("Top 20 AMG genes (by total TPM)")

# log10 transformation
gene_log_matrix_total <- gene_top20_total %>%
  dplyr::select(gene_symbol, all_of(available_samples)) %>%
  column_to_rownames("gene_symbol") %>%
  mutate(across(everything(), ~log10(.x + 1)))

gene_log_matrix_total <- gene_log_matrix_total[, available_samples]

# Set row names to italic format
rownames(gene_log_matrix_total) <- paste0("italic('", rownames(gene_log_matrix_total), "')")

# Color scheme
my_colors <- shared_heatmap_colors

# Plot pheatmap
p_amg_total <- pheatmap::pheatmap(
  gene_log_matrix_total,
  color = my_colors,
  breaks = shared_heatmap_breaks,
  legend_breaks = 0:6,
  legend_labels = as.character(0:6),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_colnames = TRUE,
  show_rownames = TRUE,
  labels_row = parse(text = rownames(gene_log_matrix_total)),
  main = "",
  fontsize_row = 10,
  fontsize_col = 18,
  fontsize = 13,
  fontfamily = "Times",
  angle_col = 0,
  cellwidth = 35,
  cellheight = 12,
  border_color = "gray80"
)

# Save figures
ggsave(path_target("Fig4b_top20_AMG_heatmap.png"),
       plot = p_amg_total, width = 8, height = 10, dpi = 300)
ggsave(path_target("Fig4b_top20_AMG_heatmap.pdf"),
       plot = p_amg_total, width = 8, height = 10)

# Save data
write.csv(gene_log_matrix_total %>% rownames_to_column("Gene"),
          path_target("top20_AMG_total_log10_TPM_matrix.csv"), row.names = FALSE)
write.xlsx(gene_top20_total, path_target("top20_AMG_total_corrected.xlsx"))

message("✅ Task 7 completed")


```

### Export Figure 4 panels as TIFF

```{r}
#| label: export-fig4-tiff
#| include: false
#| message: true
#| warning: false

tiff_dir <- path_target("tiff_panels")

dir.create(
  tiff_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# Confirm that the two plot objects were created
stopifnot(
  exists("ht"),
  exists("p_amg_total")
)

# Fig. 4a
# Existing high-resolution PNG is 3600 × 4800 pixels at 300 dpi,
# corresponding to 12 × 16 inches.
grDevices::tiff(
  filename = file.path(
    tiff_dir,
    "Fig4a_candidate_pathway_heatmap.tiff"
  ),
  width = 12,
  height = 16,
  units = "in",
  res = 300,
  compression = "lzw",
  bg = "white"
)

ComplexHeatmap::draw(ht)
grDevices::dev.off()

# Fig. 4b
# Retain the original 8 × 10 inch dimensions.
ggplot2::ggsave(
  filename = file.path(
    tiff_dir,
    "Fig4b_top20_candidate_AMG_heatmap.tiff"
  ),
  plot = p_amg_total,
  device = "tiff",
  width = 8,
  height = 10,
  units = "in",
  dpi = 300,
  compression = "lzw",
  bg = "white"
)

message(
  "Figure 4 TIFF panels exported to: ",
  normalizePath(tiff_dir)
)
```
