---
title: "06-figS6-amg-DA"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "06-figS6-amg-DA"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is ...

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "06-figS6-amg-DA")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "960c742d-536e-4c4e-a43b-c10d158fd2e5")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(data.table)
  #devtools::load_all()
})
```

## Tasks

This section is designated for your workflow implementation. **Please write all your analysis and processing code here.**

Maintaining your code within this section ensures a clean project structure and facilitates reproducibility. We recommend organizing your workflow into modular tasks—such as data cleaning, analysis, and visualization—using distinct code chunks.

Below is an example structure to get you started:

### Task 1: Data Cleaning

```{r}
#| label: data-cleaning

# Your data cleaning code here
```

### Task 2: Data Analysis

```{r}
#| label: data-analysis

# Your analysis code here
```

### Task 3: Data Visualization

```{r}
#| label: data-visualization

# Your visualization code here
```


## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r}
#| label: list-files-target
#| include: false
projthis::proj_dir_info(path_target(), tz = "CET")
```
