---
title: "10-figs7-folA-candidate"
date: today
params:
  name: "10-figs7-folA-candidate"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "10-figs7-folA-candidate")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "f71c0880-fe0b-4373-9cf3-a35ed8f87ae0")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
#| warning: false
suppressPackageStartupMessages({
  library(here)
  library(conflicted)
  library(tidyverse)
  library(data.table)
  devtools::load_all()
})
```

## Tasks

### Task 1: Read the archived Figure S7 inputs

Panel a is redrawn from the per-ORF coordinate/annotation table.
Panel b is assembled from the archived PyMOL structural overlay and TM-align
results. This QMD does not rerun structure prediction or structural alignment.
No completed panel TIFF or PNG is used as input.

```{r}
#| label: data-import
#| cache: false
#| message: false
#| warning: false

# Namespace calls below avoid conflicts with the packages already loaded.
needed <- c("png", "ragg")
missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Please install these R packages first: ", paste(missing, collapse = ", "))
}

# Historical evidence archive, not a new upstream step in the numbered workflow.
evidence_dir <- here::here("data", "16-amg-targeted-validation")
orf_file <- file.path(evidence_dir, "tables", "genomic_context_orfs.csv")
score_file <- file.path(evidence_dir, "tables", "superposition_verification.csv")
if (!all(file.exists(c(orf_file, score_file)))) {
  stop("Missing S7 input tables in: ", evidence_dir)
}

orf_all <- utils::read.csv(orf_file, stringsAsFactors = FALSE, check.names = FALSE)
orfs <- orf_all[orf_all$vOTU_id == "vOTU0027" &
                 orf_all$contig_id == "FCH3CGLCCXYL4FAA_NODE_14", , drop = FALSE]
orfs <- orfs[order(orfs$start_position), , drop = FALSE]

scores <- utils::read.csv(score_file, stringsAsFactors = FALSE, check.names = FALSE)
metrics <- scores[scores$candidate == "FCH3CGLCCXYL4FAA_NODE_14_9_esmfold" &
                    scores$reference == "1RX2", , drop = FALSE]

stopifnot(
  nrow(orfs) == 41L,
  all(orfs$contig_length_bp == 40330L),
  identical(as.integer(orfs$gene_position), 1:41),
  all(orfs$start_position >= 1 & orfs$end_position <= 40330),
  all(orfs$start_position < orfs$end_position),
  all(orfs$strand %in% c("+", "-")),
  nrow(metrics) == 1L
)
stopifnot(
  orfs$start_position[orfs$gene_position == 9] == 7392,
  orfs$end_position[orfs$gene_position == 9] == 7904,
  abs(metrics$tm_score_norm_reference - 0.82818) < 1e-6,
  abs(metrics$tm_score_norm_query - 0.77914) < 1e-6,
  abs(metrics$rmsd_reported_by_tmalign_A - 2.10) < 1e-6,
  metrics$aligned_length == 149,
  metrics$candidate_n_residues == 170,
  metrics$reference_n_residues == 159,
  tolower(metrics$pymol_align_super_cealign_used) == "false"
)

structure_file <- file.path(evidence_dir, metrics$render_png)
if (!file.exists(structure_file)) stop("Missing structural overlay: ", structure_file)
structure_image <- png::readPNG(structure_file)

# Fixed physical panel sizes; panel letters can be added in Inkscape.
panel_width <- 4.70
panel_height <- 3.35
```

### Task 2: Define the two panel drawing functions

```{r}
#| label: s7-drawing-functions
#| cache: false

draw_s7a <- function(orfs) {
  colours <- c(
    viral_structural = "#2C7FB8", viral_replication = "#7B3294",
    metabolic_candidate = "#E66101", hypothetical = "#D9D9D9"
  )
  stopifnot(all(orfs$evidence_class %in% names(colours)))
  grid::pushViewport(grid::viewport(gp = grid::gpar(fontfamily = "Times New Roman")))
  on.exit(grid::popViewport())
  grid::grid.text("vOTU0027 genomic context", y = .985, just = "top",
                  gp = grid::gpar(fontsize = 9))
  grid::grid.text("40.3 kb; CheckV low quality", y = .925, just = "top",
                  gp = grid::gpar(fontsize = 7.5))

  # Sparse labels match the current figure; all 41 ORFs are drawn.
  labels <- c("2" = "helicase", "20" = "helicase", "21" = "primase",
              "25" = "DNA pol", "40" = "minor tail")
  contig_length <- unique(orfs$contig_length_bp)
  span <- ceiling(contig_length / 3)
  track_height <- (.845 - .265) / (3 + 2 * .95)
  for (track in 0:2) {
    left <- track * span
    right <- min((track + 1) * span, contig_length)
    bottom <- .845 - track_height - track * track_height * 1.95
    grid::pushViewport(grid::viewport(
      x = .062, y = bottom, width = .926, height = track_height,
      just = c("left", "bottom"), xscale = c(left - .012 * span,
      left + 1.012 * span), yscale = c(-2.30, 2.55), clip = "off"
    ))
    grid::grid.lines(x = grid::unit(c(left, right), "native"),
                     y = grid::unit(c(0, 0), "native"),
                     gp = grid::gpar(col = "#4D4D4D", lwd = .7))
    rows <- which(orfs$end_position >= left & orfs$start_position <= right)

    # Clip full-coordinate arrows at row breaks; do not add false arrowheads.
    grid::pushViewport(grid::viewport(
      xscale = c(left - .012 * span, left + 1.012 * span),
      yscale = c(-2.30, 2.55), clip = "on"
    ))
    for (j in rows) {
      g <- orfs[j, ]
      start <- g$start_position
      end <- g$end_position
      head <- min((end - start) * .42, span * .0075)
      if (g$strand == "+") {
        x <- c(start, end - head, end - head, end, end - head,
               end - head, start)
      } else {
        x <- c(end, start + head, start + head, start, start + head,
               start + head, end)
      }
      grid::grid.polygon(
        x = grid::unit(x, "native"),
        y = grid::unit(c(-.31, -.31, -.51, 0, .51, .31, .31), "native"),
        gp = grid::gpar(fill = colours[g$evidence_class],
                        col = if (g$gene_position == 9) "black" else "#5A5A5A",
                        lwd = if (g$gene_position == 9) 1.4 else .45)
      )
    }
    grid::popViewport()

    for (j in rows) {
      g <- orfs[j, ]
      id <- as.character(g$gene_position)
      mid <- (g$start_position + g$end_position) / 2
      if (!(id %in% c(names(labels), "9")) || mid < left || mid > right) next
      grid::grid.lines(x = grid::unit(c(mid, mid), "native"),
                        y = grid::unit(c(.55, 1.04), "native"),
                        gp = grid::gpar(col = "#777777", lwd = .5))
      if (id == "9") {
        grid::grid.text(expression(paste("(annotated ", italic(folA), ")")),
          x = grid::unit(mid, "native"), y = grid::unit(1.12, "native"),
          just = "bottom", gp = grid::gpar(fontsize = 7.5))
        grid::grid.text("DHFR", x = grid::unit(mid, "native"),
          y = grid::unit(1.12, "native") + grid::unit(8, "pt"),
          just = "bottom", gp = grid::gpar(fontsize = 7.5, fontface = "bold"))
      } else {
        grid::grid.text(labels[id], x = grid::unit(mid, "native"),
          y = grid::unit(1.12, "native"), just = "bottom",
          gp = grid::gpar(fontsize = 7))
      }
      grid::grid.text(id, x = grid::unit(mid, "native"),
                      y = grid::unit(-.63, "native"), just = "top",
                      gp = grid::gpar(fontsize = 7))
    }

    ticks <- seq(0, contig_length, by = 5000)
    ticks <- ticks[ticks >= left & ticks <= right]
    grid::grid.segments(
      x0 = grid::unit(ticks, "native"), x1 = grid::unit(ticks, "native"),
      y0 = grid::unit(-2.30, "native"), y1 = grid::unit(-2.30, "native") -
        grid::unit(2.2, "pt"), gp = grid::gpar(lwd = .6)
    )
    grid::grid.text(ticks / 1000, x = grid::unit(ticks, "native"),
                    y = grid::unit(-2.30, "native") - grid::unit(3.4, "pt"),
                    just = "top", gp = grid::gpar(fontsize = 7))
    for (cap in c(0, contig_length)) {
      if (cap < left || cap > right) next
      grid::grid.lines(x = grid::unit(c(cap, cap), "native"),
                       y = grid::unit(c(-.55, .55), "native"),
                       gp = grid::gpar(lwd = 1.2))
      grid::grid.text(if (cap == 0) "contig start" else "contig end",
        x = grid::unit(cap, "native"), y = grid::unit(-1.42, "native"),
        just = c(if (cap == 0) "left" else "right", "top"),
        gp = grid::gpar(fontsize = 7))
    }
    grid::popViewport()
  }
  grid::grid.text("position on representative contig (kb)", x = .525, y = .19,
                  gp = grid::gpar(fontsize = 8))
  legend_item <- function(x, y, colour, label) {
    grid::grid.rect(x = x, y = y, width = grid::unit(7, "pt"),
                    height = grid::unit(5, "pt"), just = "left",
                    gp = grid::gpar(fill = colour, col = "#5A5A5A", lwd = .5))
    grid::grid.text(label, x = grid::unit(x, "npc") + grid::unit(9, "pt"),
                    y = y, just = "left", gp = grid::gpar(fontsize = 7.5))
  }
  legend_item(.16, .085, colours[1], "viral structural")
  legend_item(.16, .040, colours[2], "viral replication")
  legend_item(.385, .085, colours[3], "automated metabolic annotation")
  legend_item(.785, .085, colours[4], "hypothetical")
}

draw_s7b <- function(structure_image, metrics) {
  grid::pushViewport(grid::viewport(gp = grid::gpar(fontfamily = "Times New Roman")))
  on.exit(grid::popViewport())
  grid::grid.text(
    expression(paste("vOTU0027 ", italic(folA),
                     ": DHFR-family assignment supported")),
    y = .985, just = "top", gp = grid::gpar(fontsize = 9)
  )
  grid::grid.text(sprintf(
    "TM-score = %.3f (reference-normalized); RMSD = %.2f Å",
    metrics$tm_score_norm_reference, metrics$rmsd_reported_by_tmalign_A
  ), y = .925, just = "top", gp = grid::gpar(fontsize = 8))

  # Keep the structural overlay's aspect ratio and original pixels.
  grid::pushViewport(grid::viewport(x = .5, y = .495, width = .96, height = .70))
  available_w <- grid::convertWidth(grid::unit(1, "npc"), "in", valueOnly = TRUE)
  available_h <- grid::convertHeight(grid::unit(1, "npc"), "in", valueOnly = TRUE)
  image_ratio <- dim(structure_image)[2] / dim(structure_image)[1]
  image_h <- min(available_h, available_w / image_ratio)
  grid::grid.raster(structure_image, width = grid::unit(image_h * image_ratio, "in"),
                    height = grid::unit(image_h, "in"), interpolate = TRUE)
  grid::popViewport()
  legend_item <- function(x, y, colour, label) {
    grid::grid.rect(x = x, y = y, width = grid::unit(7, "pt"),
                    height = grid::unit(5, "pt"), just = "left",
                    gp = grid::gpar(fill = colour, col = "black", lwd = .5))
    grid::grid.text(label, x = grid::unit(x, "npc") + grid::unit(9, "pt"),
                    y = y, just = "left", gp = grid::gpar(fontsize = 7.5))
  }
  legend_item(.23, .078, "#E66101", "candidate, superposed")
  legend_item(.23, .034, "#F9CFA6", "candidate, not superposed")
  legend_item(.57, .078, "#2C7FB8", "reference (PDB 1RX2)")
}

```

### Task 3: Export the two panels

All final TIFF and preview exports are kept together in this chunk.

```{r}
#| label: s7-export-panels
#| cache: false
#| message: false
#| warning: false

tiff_dir <- path_target("tiff_panels")
preview_dir <- path_target("preview_png")
dir.create(tiff_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(preview_dir, recursive = TRUE, showWarnings = FALSE)

save_panel <- function(draw_function, stem) {
  write_one <- function(filename, dpi, is_tiff) {
    if (is_tiff) {
      ragg::agg_tiff(filename, width = panel_width, height = panel_height,
                     units = "in", res = dpi, background = "white",
                     compression = "lzw", bitsize = 8)
    } else {
      ragg::agg_png(filename, width = panel_width, height = panel_height,
                    units = "in", res = dpi, background = "white")
    }
    on.exit(grDevices::dev.off(), add = TRUE)
    grid::grid.newpage()
    draw_function()
  }
  write_one(file.path(tiff_dir, paste0(stem, ".tiff")), 600, TRUE)
  write_one(file.path(preview_dir, paste0(stem, ".png")), 200, FALSE)
}

save_panel(function() draw_s7a(orfs), "FigS7a_vOTU0027_genomic_context")
save_panel(function() draw_s7b(structure_image, metrics), "FigS7b_folA_structure")

# Small audit outputs: the exact records and inputs used for these panels.
utils::write.csv(orfs, path_target("FigS7a_ORF_coordinates_used.csv"), row.names = FALSE)
utils::write.csv(metrics, path_target("FigS7b_structure_metrics_used.csv"), row.names = FALSE)
input_files <- c(orf_file, score_file, structure_file)
utils::write.csv(
  data.frame(input = input_files, md5 = unname(tools::md5sum(input_files))),
  path_target("FigS7_input_manifest.csv"), row.names = FALSE
)
writeLines(capture.output(sessionInfo()), path_target("FigS7_sessionInfo.txt"))
```

### Preview

```{r}
#| label: s7-preview
#| echo: false
#| out-width: "100%"
knitr::include_graphics(c(
  file.path(preview_dir, "FigS7a_vOTU0027_genomic_context.png"),
  file.path(preview_dir, "FigS7b_folA_structure.png")
))
```

## Files written

These files have been written to the target directory, `r paste0("data/", params$name)`:

```{r}
#| label: list-files-target
projthis::proj_dir_info(path_target(), tz = "CET") %>% knitr::kable()
```

