---
title: "10-figs5-rare-patterns"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "10-figs5-rare-patterns"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is ...

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "10-figs5-rare-patterns")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "150b48a9-4c94-4c0f-810a-0226fa12ae65")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(TreeSummarizedExperiment)
  library(ggplot2)
  library(patchwork)
  library(scales)
})
```

## Tasks

This section is designated for your workflow implementation. **Please write all your analysis and processing code here.**

Maintaining your code within this section ensures a clean project structure and facilitates reproducibility. We recommend organizing your workflow into modular tasks—such as data cleaning, analysis, and visualization—using distinct code chunks.

Below is an example structure to get you started:

### Task 1: Load TSE and Extract Metadata

```{r}
#| label: load-data

tse <- readRDS(path_source("01-tse-construction", "tse.rds"))

contig_anno <- metadata(tse)$contig_annotation
imgvr <- metadata(tse)$imgvr_source

message("TSE loaded: ", ncol(tse), " samples, ", nrow(tse), " vOTUs")
```

### Task 2: Figure S5a - Bubble Plot (Rare)

```{r}
#| label: figs5a-bubble
#| fig.width: 8.5
#| fig.height: 6

tpm_mat <- assays(tse)$tpm

family_df <- data.frame(
  vOTU_id = rownames(tse),
  family = as.character(rowData(tse)$family),
  stringsAsFactors = FALSE
)

sample_df <- data.frame(
  sample_id = colnames(tse),
  sample_group = as.character(colData(tse)$sample_group),
  stringsAsFactors = FALSE
)

abundance_long <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "tpm") %>%
  dplyr::left_join(sample_df, by = "sample_id") %>%
  dplyr::left_join(family_df, by = "vOTU_id")

abundance_long <- abundance_long %>%
  mutate(
    family = str_remove(family, regex("^(\\[?Family\\]?_?|f__|family_)", ignore_case = TRUE)),
    family = str_trim(family),
    family = if_else(is.na(family) | family == "", "Unclassified virus", family)
  )

sample_totals <- abundance_long %>%
  group_by(sample_group) %>%
  summarise(total_TPM = sum(tpm, na.rm = TRUE), .groups = "drop")

abundance_rel <- abundance_long %>%
  dplyr::left_join(sample_totals, by = "sample_group") %>%
  mutate(rel_abundance = tpm / total_TPM)

# Identify RARE vOTUs (< 0.1% = < 0.001)
rare_vOTUs <- abundance_rel %>%
  dplyr::filter(rel_abundance < 0.001 & rel_abundance > 0) %>%
  dplyr::select(sample_group, vOTU_id, family, rel_abundance) %>%
  arrange(sample_group, desc(rel_abundance))

df_sum <- rare_vOTUs %>%
  group_by(family, sample_group) %>%
  summarise(total_abund = sum(rel_abundance), .groups = "drop")

top_families <- df_sum %>%
  group_by(family) %>%
  summarise(mean_abund = mean(total_abund), .groups = "drop") %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 14) %>%
  pull(family)

df_sum <- df_sum %>%
  dplyr::filter(family %in% top_families) %>%
  mutate(
    sample_group = factor(sample_group, levels = c("BS", "SA", "IA", "DA")),
    family = forcats::fct_relevel(family, "Unclassified virus", after = 0)
  )

p_bubble <- ggplot(df_sum, aes(x = total_abund, y = forcats::fct_rev(factor(family)), color = sample_group)) +
  geom_point(aes(size = total_abund), alpha = 0.85) +
  scale_size_continuous(
    range = c(2, 10),
    name = "Relative Abundance",
    breaks = pretty(df_sum$total_abund, n = 4)
  ) +
  scale_x_continuous(
    breaks = seq(0, 0.8, by = 0.2),
    limits = c(0, max(df_sum$total_abund) * 1.1),
    labels = number_format(accuracy = 0.1)
  ) +
  scale_color_manual(
    values = c("BS" = "#E69F00", "SA" = "#56B4E9", "IA" = "#009E73", "DA" = "#F0E442"),
    name = "Sample"
  ) +
  labs(x = "Relative abundance", y = NULL) +
  theme_minimal(base_size = 22) +
  theme(
    text = element_text(family = "Times"),
    axis.line = element_line(linewidth = 0.8, color = "black"),
    axis.ticks = element_line(linewidth = 0.5, color = "black"),
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "italic", size = 18, color = "black"),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "plain", color = "black"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p_bubble)
ggsave(path_target("FigS5a_bubble.png"), p_bubble, width = 8.5, height = 6, dpi = 300)
write_csv(df_sum, path_target("FigS5a_bubble_data.csv"))
write_csv(rare_vOTUs, path_target("FigS5a_rare_full.csv"))

message("FigS5a completed: ", nrow(rare_vOTUs), " rare vOTU-sample pairs")
```

### Task 3: Figure S5b - Ecosystem Distribution 
```{r}
#| label: figs5b-ecosystem
#| fig.width: 8
#| fig.height: 6

tpm_mat <- assays(tse)$tpm

sample_df <- data.frame(
  sample_id = colnames(tse),
  sample_group = as.character(colData(tse)$sample_group),
  stringsAsFactors = FALSE
)

abundance_long <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM") %>%
  dplyr::left_join(sample_df, by = "sample_id")

sample_totals <- abundance_long %>%
  group_by(sample_group) %>%
  summarise(total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop")

abundance_rel <- abundance_long %>%
  dplyr::left_join(sample_totals, by = "sample_group") %>%
  mutate(rel_abundance = TPM / total_TPM)

# Identify RARE vOTUs (< 0.1%)
rare_votus_list <- abundance_rel %>%
  dplyr::filter(rel_abundance < 0.001 & rel_abundance > 0) %>%
  pull(vOTU_id) %>%
  unique()

abundance_rare <- abundance_rel %>%
  dplyr::filter(vOTU_id %in% rare_votus_list)

# Extract ecosystem classification
extract_level5_position5 <- function(eco_string) {
  if (is.na(eco_string) || eco_string == "") return(NA)
  parts <- strsplit(as.character(eco_string), ";")[[1]]
  parts <- trimws(parts)
  if (length(parts) >= 5 && parts[5] != "") {
    return(parts[5])
  } else {
    return(NA)
  }
}

imgvr_eco <- imgvr %>%
  dplyr::select(contig_id, eco_class = `Ecosystem classification`) %>%
  mutate(level5 = sapply(eco_class, extract_level5_position5)) %>%
  dplyr::filter(!is.na(level5)) %>%
  dplyr::select(contig_id, level5) %>%
  dplyr::distinct()

# Link ecosystem to vOTUs (with weighting)
data_with_eco <- contig_anno %>%
  dplyr::left_join(imgvr_eco, by = "contig_id") %>%
  dplyr::filter(!is.na(level5))

# Calculate weights for vOTUs with multiple ecosystems
votu_eco_count <- data_with_eco %>%
  dplyr::select(vOTU_id, level5) %>%
  dplyr::distinct() %>%
  group_by(vOTU_id) %>%
  summarise(n_ecosystems = dplyr::n_distinct(level5), .groups = "drop")

data_with_eco_weighted <- data_with_eco %>%
  dplyr::select(vOTU_id, level5) %>%
  dplyr::distinct() %>%
  dplyr::left_join(votu_eco_count, by = "vOTU_id") %>%
  mutate(weight = 1 / n_ecosystems)

# Join with abundance (many-to-many expected)
abundance_rare_eco <- abundance_rare %>%
  dplyr::inner_join(data_with_eco_weighted, by = "vOTU_id", relationship = "many-to-many") %>%
  mutate(weighted_TPM = TPM * weight)

# Calculate ecosystem abundance (using weighted TPM)
abundance_by_eco_rare <- abundance_rare_eco %>%
  group_by(sample_group, level5) %>%
  summarise(
    level5_tpm = sum(weighted_TPM, na.rm = TRUE),
    n_viruses = dplyr::n_distinct(vOTU_id),
    .groups = "drop"
  )

abundance_by_eco_rare <- abundance_by_eco_rare %>%
  dplyr::left_join(sample_totals, by = "sample_group") %>%
  mutate(relative_abundance = level5_tpm / total_TPM) %>%
  dplyr::select(sample_group, level5, level5_tpm, n_viruses, relative_abundance, total_TPM)

# Define Level5 order and colors
level5_order <- c(
  "Host-associated(human)", "Host-associated(plants)", "Host-associated(other)",
  "Marine", "Freshwater", "Aquatic(sediment)", "Non-marine_Saline_and_Alkaline",
  "Terrestrial(soil)", "Terrestrial(other)", "Thermal_springs", "Engineered"
)

host_colors <- c("#F9E2C8", "#F4CCAC", "#C4714F")
water_colors <- c("#D4F0CD", "#B8E6AD", "#99DD87", "#2E7D32")
terrestrial_colors <- c("#D9C9BD", "#6D4C41")
special_colors <- c("#E8E4D8", "#5D4037")
all_colors <- c(host_colors, water_colors, terrestrial_colors, special_colors)
color_map <- setNames(all_colors, level5_order)

# Prepare plot data
plot_data <- abundance_by_eco_rare %>%
  mutate(
    sample_group = factor(sample_group, levels = c("BS", "SA", "IA", "DA")),
    level5 = factor(level5, levels = level5_order)
  ) %>%
  arrange(sample_group, desc(relative_abundance))

# Create vertical bar chart
fig <- ggplot(plot_data, aes(x = sample_group, y = relative_abundance, fill = level5)) +
  geom_col(position = "stack", width = 0.6, color = "white", linewidth = 0.2) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = function(x) sprintf("%.1f", x),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = color_map,
    name = "Level 5 Ecosystem",
    breaks = level5_order
  ) +
  labs(x = NULL, y = "Relative Abundance") +
  theme_minimal(base_size = 22) +
  theme(
    text = element_text(family = "Times"),
    axis.text = element_text(color = "black", size = 18),
    axis.title.y = element_text(size = 20, face = "plain"),
    axis.title.x = element_blank(),
    axis.line = element_line(color = "black", linewidth = 1),
    axis.ticks = element_line(color = "black", linewidth = 0.8),
    legend.position = "right",
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12, face = "bold"),
    legend.key.size = unit(0.5, "cm"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(15, 20, 10, 10)
  )

print(fig)
ggsave(path_target("FigS5b_ecosystem.png"), fig, width = 8, height = 6, dpi = 300)
write_csv(abundance_by_eco_rare, path_target("FigS5b_ecosystem_data.csv"))

message("FigS5b completed: ", length(rare_votus_list), " rare vOTUs")

```


### Task 4: Figure S5c - Host Heatmap 

```{r}
#| label: figs5c-host-heatmap
#| fig.width: 8
#| fig.height: 10

host_genome <- metadata(tse)$host_genome_edges

CONFIDENCE_THRESHOLD <- 80

host_filtered <- host_genome %>%
  dplyr::filter(Confidence.score >= CONFIDENCE_THRESHOLD) %>%
  mutate(
    Phylum = str_extract(Host.taxonomy, "(?<=p__)[^;]+"),
    Phylum = if_else(is.na(Phylum), "Unclassified", Phylum)
  )

# Use rare vOTU list (from S5a/S5b)
if (!exists("rare_votus_list")) {
  sample_df <- data.frame(
    sample_id = colnames(tse),
    sample_group = as.character(colData(tse)$sample_group),
    stringsAsFactors = FALSE
  )
  
  tpm_mat <- assays(tse)$tpm
  
  abundance_long <- as.data.frame(tpm_mat) %>%
    rownames_to_column("vOTU_id") %>%
    pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM") %>%
    dplyr::left_join(sample_df, by = "sample_id")
  
  sample_totals <- abundance_long %>%
    group_by(sample_group) %>%
    summarise(total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop")
  
  abundance_rel <- abundance_long %>%
    dplyr::left_join(sample_totals, by = "sample_group") %>%
    mutate(rel_abundance = TPM / total_TPM)
  
  rare_votus_list <- abundance_rel %>%
    dplyr::filter(rel_abundance < 0.001 & rel_abundance > 0) %>%
    pull(vOTU_id) %>%
    unique()
}

# Filter host predictions for rare vOTUs only
host_rare <- host_filtered %>%
  dplyr::filter(vOTU_id %in% rare_votus_list)

# ============================================================================
# NEW: Calculate weights for vOTUs with multiple Phylum predictions
# ============================================================================

# Step 1: Count how many different Phyla each vOTU has
host_count <- host_rare %>%
  dplyr::select(vOTU_id, Phylum) %>%
  dplyr::distinct() %>%
  group_by(vOTU_id) %>%
  summarise(n_hosts = dplyr::n_distinct(Phylum), .groups = "drop")

# Step 2: Calculate weights (1 / n_hosts)
host_weighted <- host_rare %>%
  dplyr::select(vOTU_id, Phylum) %>%
  dplyr::distinct() %>%
  dplyr::left_join(host_count, by = "vOTU_id") %>%
  mutate(weight = 1 / n_hosts)

# Diagnostic: Check multi-Phylum vOTUs
multi_phylum_votus <- host_count %>% dplyr::filter(n_hosts > 1)
if (nrow(multi_phylum_votus) > 0) {
  message(sprintf("  %d rare vOTUs have predictions for multiple Phyla (weighted)", nrow(multi_phylum_votus)))
} else {
  message("  All rare vOTUs have predictions for only one Phylum")
}

# ============================================================================
# Prepare abundance data
# ============================================================================

sample_df <- data.frame(
  sample_id = colnames(tse),
  sample_group = as.character(colData(tse)$sample_group),
  stringsAsFactors = FALSE
)

tpm_rare <- as.data.frame(assays(tse)$tpm) %>%
  rownames_to_column("vOTU_id") %>%
  dplyr::filter(vOTU_id %in% rare_votus_list) %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "tpm") %>%
  dplyr::left_join(sample_df, by = "sample_id") %>%
  dplyr::filter(!is.na(sample_group))

# ============================================================================
# Create heatmap data (using WEIGHTED vOTU counts)
# ============================================================================

# Join with weighted host predictions
host_abundance <- host_weighted %>%  # ← Changed: use host_weighted
  dplyr::left_join(tpm_rare, by = "vOTU_id", relationship = "many-to-many")

# Calculate weighted vOTU counts per Phylum
heatmap_data <- host_abundance %>%
  dplyr::filter(tpm > 0) %>%
  group_by(sample_group, Phylum) %>%
  summarise(
    n_vOTUs_weighted = sum(weight),  # ← Changed: use weighted sum
    presence = if_else(sum(weight) > 0, 1, 0),  # ← Changed: based on weighted sum
    .groups = "drop"
  )

# Complete the grid (only rare vOTUs' Phyla)
all_samples <- c("BS", "SA", "IA", "DA")
all_phyla <- unique(host_weighted$Phylum)  # ← Changed: use host_weighted

heatmap_data_complete <- expand.grid(
  sample_group = all_samples,
  Phylum = all_phyla,
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(heatmap_data, by = c("sample_group", "Phylum")) %>%
  mutate(
    n_vOTUs_weighted = if_else(is.na(n_vOTUs_weighted), 0, n_vOTUs_weighted),  # ← Changed
    presence = if_else(is.na(presence), 0, presence)
  )

# Order phyla by total presence
phylum_order <- heatmap_data_complete %>%
  group_by(Phylum) %>%
  summarise(total_presence = sum(presence), .groups = "drop") %>%
  arrange(desc(total_presence)) %>%
  pull(Phylum)

heatmap_data_complete <- heatmap_data_complete %>%
  mutate(
    sample_group = factor(sample_group, levels = c("BS", "SA", "IA", "DA")),
    Phylum = factor(Phylum, levels = phylum_order)
  )

# Create compact heatmap (horizontal layout)
p_heatmap <- ggplot(heatmap_data_complete, aes(x = sample_group, y = Phylum)) +
  geom_tile(aes(fill = factor(presence)), color = "white", linewidth = 1) +
  scale_fill_manual(
    values = c("0" = "gray90", "1" = "#66C2A5"),
    guide = "none"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    text = element_text(family = "Times"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 18, color = "black"),
    axis.text.y = element_text(color = "black", size = 18, face = "italic"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p_heatmap)

ggsave(path_target("FigS5c_host_heatmap.png"), p_heatmap, width = 8, height = 10, dpi = 300)
write_csv(heatmap_data_complete, path_target("FigS5c_host_heatmap_data.csv"))

summary_data <- heatmap_data_complete %>%
  group_by(sample_group) %>%
  summarise(
    n_phyla_present = sum(presence > 0),
    n_vOTUs_weighted = sum(n_vOTUs_weighted),  # ← Changed: use weighted sum
    .groups = "drop"
  )

write_csv(summary_data, path_target("FigS5c_summary.csv"))

message("FigS5c completed: ", dplyr::n_distinct(host_weighted$vOTU_id), 
        " rare vOTUs with host predictions (weighted)")

```



