---
title: "10-figs5-rare-patterns"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "10-figs5-rare-patterns"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is ...

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "10-figs5-rare-patterns")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "150b48a9-4c94-4c0f-810a-0226fa12ae65")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: packages
#| message: false
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(TreeSummarizedExperiment)
  library(ggplot2)
  library(patchwork)
  library(scales)
})
```

## Tasks

This section is designated for your workflow implementation. **Please write all your analysis and processing code here.**

Maintaining your code within this section ensures a clean project structure and facilitates reproducibility. We recommend organizing your workflow into modular tasks—such as data cleaning, analysis, and visualization—using distinct code chunks.

Below is an example structure to get you started:

### Task 1: Load TSE and Extract Metadata

```{r}
#| label: load-data
#| message: true

message("========================================================================")
message("Loading TSE and extracting metadata...")
message("========================================================================\n")

# Load TSE
tse <- readRDS(path_source("01-tse-construction", "tse.rds"))
message("✅ TSE loaded: ", ncol(tse), " samples, ", nrow(tse), " vOTUs\n")

# Extract annotation data from TSE metadata
contig_anno <- metadata(tse)$contig_annotation
imgvr <- metadata(tse)$imgvr_source
stopifnot(!is.null(contig_anno), !is.null(imgvr))

message("✅ contig_annotation extracted: ", nrow(contig_anno), " contigs")
message("✅ imgvr_source extracted: ", nrow(imgvr), " entries\n")

```

### Task 2: Figure S5a: Rare Viral Taxa (Family Level Bubble Plot)

```{r}
#| label: figs5a-bubble
#| message: true
#| fig.width: 8.5
#| fig.height: 6

message("========================================================================")
message("Figure S5a: Rare Viral Taxa Bubble Plot (Family Level)")
message("Criterion: Relative Abundance < 0.1% per sample group")
message("========================================================================\n")

# ============================================================================
# STEP 1: Extract data from TSE
# ============================================================================
message("📊 Extracting data from TSE...")

# Extract TPM assay
tpm_mat <- assays(tse)$tpm

# Extract family from rowData
family_df <- data.frame(
  vOTU_id = rownames(tse),
  family = as.character(rowData(tse)$family),
  stringsAsFactors = FALSE
)

# Extract sample_group from colData
sample_df <- data.frame(
  sample_id = colnames(tse),
  sample_source1 = as.character(colData(tse)$sample_group),
  stringsAsFactors = FALSE
)

message("✅ Data extracted from TSE")
cat("Samples:", nrow(sample_df), "\n")
cat("Sample groups:", paste(unique(sample_df$sample_source1), collapse = ", "), "\n\n")

# ============================================================================
# STEP 2: Identify RARE vOTUs (< 0.1% and > 0, per sample_group)
# ============================================================================
message("🔍 Calculating relative abundance (per sample_group)...")

# Convert to long format
abundance_long <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "tpm") %>%
  dplyr::left_join(sample_df, by = "sample_id") %>%
  dplyr::left_join(family_df, by = "vOTU_id")

# Clean family names
abundance_long <- abundance_long %>%
  mutate(
    family = str_remove(family, regex("^(\\[?Family\\]?_?|f__|family_)", ignore_case = TRUE)),
    family = str_trim(family),
    family = ifelse(is.na(family) | family == "", "Unclassified virus", family)
  )

# Calculate total TPM per sample_group
sample_totals <- abundance_long %>%
  group_by(sample_source1) %>%
  summarise(total_TPM = sum(tpm, na.rm = TRUE), .groups = "drop")

# Calculate relative abundance (decimal: 0-1)
abundance_rel <- abundance_long %>%
  dplyr::left_join(sample_totals, by = "sample_source1") %>%
  mutate(rel_abundance = tpm / total_TPM)

# *** Identify RARE vOTUs (< 0.1% = < 0.001 AND > 0 per sample_group) ***
rare_vOTUs <- abundance_rel %>%
  filter(rel_abundance < 0.001 & rel_abundance > 0) %>%
  select(sample_source1, vOTU_id, family, rel_abundance) %>%
  arrange(sample_source1, desc(rel_abundance))

message(sprintf("✅ Identified %d rare vOTU-sample pairs\n", nrow(rare_vOTUs)))

# ============================================================================
# STEP 3: Prepare plot data
# ============================================================================
message("📊 Preparing plot data...")

# Sum abundance per family per sample_group
df_sum <- rare_vOTUs %>%
  group_by(family, sample_source1) %>%
  summarise(total_abund = sum(rel_abundance), .groups = "drop")

# Keep top 14 families
top_families <- df_sum %>%
  group_by(family) %>%
  summarise(mean_abund = mean(total_abund), .groups = "drop") %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 14) %>%
  pull(family)

df_sum <- df_sum %>%
  filter(family %in% top_families) %>%
  mutate(
    sample_source1 = factor(sample_source1, levels = c("BS", "SA", "IA", "DA")),
    family = fct_relevel(family, "Unclassified virus", after = 0)
  )

message("✅ Plot data ready\n")

# ============================================================================
# STEP 4: Create bubble plot
# ============================================================================
message("🎨 Creating bubble plot...")

p_bubble <- ggplot(df_sum, aes(x = total_abund, y = fct_rev(factor(family)), color = sample_source1)) +
  geom_point(aes(size = total_abund), alpha = 0.85) +
  scale_size_continuous(
    range = c(2, 10),
    name = "Relative Abundance",
    breaks = pretty(df_sum$total_abund, n = 4)
  ) +
  scale_x_continuous(
    breaks = seq(0, 0.8, by = 0.2),
    limits = c(0, max(df_sum$total_abund) * 1.1),
    labels = number_format(accuracy = 0.1)
  ) +
  scale_color_manual(
    values = c("BS" = "#E69F00", "SA" = "#56B4E9", "IA" = "#009E73", "DA" = "#F0E442"),
    name = "Sample"
  ) +
  labs(x = "Relative abundance", y = NULL) +
  theme_minimal(base_size = 22) +
  theme(
    text = element_text(family = "Times"),
    axis.line.x = element_line(linewidth = 0.8, color = "black"),
    axis.line.y = element_line(linewidth = 0.8, color = "black"),
    axis.ticks.x = element_line(linewidth = 0.5, color = "black"),
    axis.ticks.y = element_line(linewidth = 0.5, color = "black"),
    panel.grid = element_blank(),
    axis.text.y = element_text(face = "italic", size = 18, family = "Times", color = "black"),
    axis.text.x = element_text(size = 18, family = "Times", color = "black"),
    axis.title.x = element_text(size = 20, face = "plain", family = "Times", color = "black"),
    legend.title = element_text(size = 12, face = "bold", family = "Times"),
    legend.text = element_text(size = 11, family = "Times"),
    legend.background = element_blank(),
    legend.key = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p_bubble)

# Save
ggsave(path_target("FigS5a_bubble.pdf"), plot = p_bubble, width = 8.5, height = 6, dpi = 300)
ggsave(path_target("FigS5a_bubble.png"), plot = p_bubble, width = 8.5, height = 6, dpi = 300)
write_csv(df_sum, path_target("FigS5a_bubble_data.csv"))
write_csv(rare_vOTUs, path_target("FigS5a_rare_full.csv"))

message("✅ Figure S5a completed!\n")

```

### Task 3: Figure S5b: Rare Viral Groups - Ecosystem Distribution
```{r}
#| label: figs5b-ecosystem
#| message: true
#| fig.width: 8
#| fig.height: 6

message("========================================================================")
message("Figure S5b: Rare Viral Groups - Ecosystem Distribution")
message("Criterion: Relative Abundance < 0.1% per sample group")
message("Denominator: Total TPM of ALL viruses in each sample group")
message("========================================================================\n")

# ============================================================================
# STEP 1: Extract TPM and calculate relative abundance (per sample_group)
# ============================================================================
message("📊 Calculating relative abundance...")

# Extract TPM matrix from TSE
tpm_mat <- assays(tse)$tpm

# Create sample mapping using TSE colData
sample_df <- data.frame(
  sample_id = colnames(tse),
  sample_group = as.character(colData(tse)$sample_group),
  stringsAsFactors = FALSE
)

# Convert to long format
abundance_long <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM") %>%
  dplyr::left_join(sample_df, by = "sample_id")

# Calculate total TPM per sample_group (all viruses)
sample_totals <- abundance_long %>%
  group_by(sample_group) %>%
  summarise(total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop")

# Calculate relative abundance
abundance_rel <- abundance_long %>%
  dplyr::left_join(sample_totals, by = "sample_group") %>%
  mutate(rel_abundance = TPM / total_TPM)

# Identify RARE vOTUs (< 0.1% and > 0)
rare_votus_list <- abundance_rel %>%
  filter(rel_abundance < 0.001 & rel_abundance > 0) %>%
  pull(vOTU_id) %>%
  unique()

message(sprintf("✅ Identified %d rare vOTUs (< 0.1%%)\n", length(rare_votus_list)))

# Filter rare vOTUs
abundance_rare <- abundance_rel %>%
  filter(vOTU_id %in% rare_votus_list)

# ============================================================================
# STEP 2: Extract ecosystem classification (from TSE metadata)
# ============================================================================
message("🌍 Extracting ecosystem classification...")

# Function to extract Level5 (exactly as old code)
extract_level5_position5 <- function(eco_string) {
  if (is.na(eco_string) || eco_string == "") return(NA)
  parts <- strsplit(as.character(eco_string), ";")
  parts <- parts[[1]]
  parts <- trimws(parts)
  if (length(parts) >= 5 && parts[5] != "") {
    return(parts[5])
  } else {
    return(NA)
  }
}

# Extract Level5 from imgvr (now from TSE metadata)
imgvr_eco <- imgvr %>%
  select(contig_id, eco_class = `Ecosystem classification`) %>%
  mutate(level5 = sapply(eco_class, extract_level5_position5)) %>%
  filter(!is.na(level5)) %>%
  select(contig_id, level5) %>%
  distinct()

message("✅ Ecosystem data extracted\n")


# ============================================================================
# STEP 3: Link ecosystem to vOTUs (using ALL contigs from metadata)
# ============================================================================
message("🔗 Linking ecosystem classification (using ALL contigs)...")

# Use ALL contigs (from TSE metadata)
data_with_eco <- contig_anno %>%
  dplyr::left_join(imgvr_eco, by = "contig_id") %>%
  filter(!is.na(level5))

# Combine with abundance data (fix many-to-many warning)
abundance_rare_eco <- abundance_rare %>%
  filter(vOTU_id %in% data_with_eco$vOTU_id) %>%
  dplyr::left_join(
    data_with_eco %>% select(vOTU_id, level5) %>% distinct(), 
    by = "vOTU_id",
    relationship = "many-to-many"  # ← 添加这个来消除warning
  )

message("✅ Ecosystem data linked\n")

# ============================================================================
# STEP 4: Calculate relative abundance per ecosystem
# ============================================================================
message("📊 Calculating ecosystem abundance...")

# 注意：改成 sample_group
abundance_by_eco_rare <- abundance_rare_eco %>%
  filter(!is.na(level5)) %>%
  group_by(sample_group, level5) %>%  # ← 改成 sample_group
  summarise(
    level5_tpm = sum(TPM, na.rm = TRUE),
    n_viruses = n_distinct(vOTU_id),
    .groups = "drop"
  )

# Use total TPM of ALL viruses as denominator
abundance_by_eco_rare <- abundance_by_eco_rare %>%
  dplyr::left_join(sample_totals, by = "sample_group") %>%  # ← 改成 sample_group 和 sample_totals
  mutate(relative_abundance = level5_tpm / total_TPM) %>%  # ← 改成 total_TPM
  select(sample_group, level5, level5_tpm, n_viruses, relative_abundance, total_TPM)

message(sprintf("✅ Ecosystem data processed (%d rare vOTUs with ecosystem annotation)\n",
                length(unique(abundance_rare_eco$vOTU_id[!is.na(abundance_rare_eco$level5)]))))

# ============================================================================
# STEP 5: Define Level5 order and colors (exactly as old code)
# ============================================================================
level5_order <- c(
  "Host-associated(human)",
  "Host-associated(plants)",
  "Host-associated(other)",
  "Marine",
  "Freshwater",
  "Aquatic(sediment)",
  "Non-marine_Saline_and_Alkaline",
  "Terrestrial(soil)",
  "Terrestrial(other)",
  "Thermal_springs",
  "Engineered"
)

host_colors <- c("#F9E2C8", "#F4CCAC", "#C4714F")
water_colors <- c("#D4F0CD", "#B8E6AD", "#99DD87", "#2E7D32")
terrestrial_colors <- c("#D9C9BD", "#6D4C41")
special_colors <- c("#E8E4D8", "#5D4037")
all_colors <- c(host_colors, water_colors, terrestrial_colors, special_colors)
color_map <- setNames(all_colors, level5_order)

# ============================================================================
# STEP 6: Prepare plot data
# ============================================================================
plot_data <- abundance_by_eco_rare %>%
  mutate(sample_group = factor(sample_group, levels = c("BS", "SA", "IA", "DA"))) %>%  # ← 改成 sample_group
  mutate(level5 = factor(level5, levels = level5_order)) %>%
  arrange(sample_group, desc(relative_abundance))

# ============================================================================
# STEP 7: Create vertical bar chart (exactly matching old code)
# ============================================================================
message("🎨 Creating vertical bar chart...")

fig <- ggplot(plot_data, aes(x = sample_group, y = relative_abundance, fill = level5)) +  # ← 改成 sample_group
  geom_col(position = "stack", width = 0.6, color = "white", linewidth = 0.2) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = function(x) sprintf("%.1f", x),
    expand = c(0, 0)
  ) +
  scale_x_discrete() +
  scale_fill_manual(
    values = color_map,
    name = "Level 5 Ecosystem",
    breaks = level5_order
  ) +
  labs(
    x = NULL,
    y = "Relative Abundance"
  ) +
  theme_minimal(base_size = 22) +
  theme(
    text = element_text(family = "Times"),
    axis.text.x = element_text(color = "black", size = 18, family = "Times"),
    axis.text.y = element_text(color = "black", size = 18, family = "Times"),
    axis.title.y = element_text(size = 20, face = "plain", family = "Times"),
    axis.title.x = element_blank(),
    axis.line = element_line(color = "black", linewidth = 1),
    axis.ticks = element_line(color = "black", linewidth = 0.8),
    legend.position = "right",
    legend.text = element_text(size = 11, family = "Times"),
    legend.title = element_text(size = 12, face = "bold", family = "Times"),
    legend.key.size = unit(0.5, "cm"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(15, 20, 10, 10)
  )

print(fig)

# Save plot
ggsave(path_target("FigS5b_ecosystem.pdf"), plot = fig, width = 8, height = 6, dpi = 300)
ggsave(path_target("FigS5b_ecosystem.png"), plot = fig, width = 8, height = 6, dpi = 300)

# Save data
write_csv(abundance_by_eco_rare, path_target("FigS5b_ecosystem_data.csv"))

message("✅ Figure S5b completed and saved!\n")


```


### Figure S5c: Predicted Bacterial Host of Rare Viruses (Heatmap)

```{r}
#| label: figs5c-host-heatmap
#| message: true
#| fig.width: 8
#| fig.height: 10

message("========================================================================")
message("Figure S5c: Predicted Bacterial Host of Rare Viruses (Heatmap)")
message("Confidence threshold: ≥75")
message("========================================================================\n")

# ============================================================================
# STEP 1: Load host prediction data from TSE metadata
# ============================================================================
message("📥 Loading host prediction data from TSE metadata...")

# 正确的 metadata key（有下划线）
host_genome <- metadata(tse)$host_genome_edges  # ← 注意是 host_genome_edges

stopifnot(!is.null(host_genome))

message("✅ host_genome loaded: ", nrow(host_genome), " predictions")
message("   Columns: ", paste(colnames(host_genome), collapse = ", "), "\n")


# ============================================================================
# STEP 2: Filter high-confidence host predictions (≥75)
# ============================================================================
CONFIDENCE_THRESHOLD <- 80

message(sprintf("🔍 Filtering host predictions (confidence ≥ %d)...\n", CONFIDENCE_THRESHOLD))

host_filtered <- host_genome %>%
  filter(Confidence.score >= CONFIDENCE_THRESHOLD) %>%
  mutate(
    Phylum = str_extract(Host.taxonomy, "(?<=p__)[^;]+"),
    Phylum = if_else(is.na(Phylum), "Unclassified", Phylum)
  )

message(sprintf("✅ High-confidence predictions: %d rows", nrow(host_filtered)))
message(sprintf("✅ Unique phyla: %d\n", n_distinct(host_filtered$Phylum)))

# ============================================================================
# STEP 3: Use rare vOTU list from Fig S5a/S5b (already calculated)
# ============================================================================
message("📋 Using rare vOTU list from previous analysis...")

# rare_votus_list should already exist from Fig S5a/S5b
# If not, recalculate (fallback)
if (!exists("rare_votus_list")) {
  message("⚠️ Recalculating rare vOTUs...")
  
  sample_map <- data.frame(
    sample_id = c("FCH3CGLCCXYL2DAA", "FCH3CGLCCXYL2EAA", "FCH3CGLCCXYL4FAA", "TARATASF"),
    sample_short = c("SA", "IA", "BS", "DA"),
    stringsAsFactors = FALSE
  )
  
  tpm_mat <- assays(tse)$tpm
  
  abundance_long <- as.data.frame(tpm_mat) %>%
    rownames_to_column("vOTU_id") %>%
    pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM") %>%
    dplyr::left_join(sample_map, by = "sample_id")
  
  sample_total_tpm_all <- abundance_long %>%
    group_by(sample_short) %>%
    summarise(total_tpm_all = sum(TPM, na.rm = TRUE), .groups = "drop")
  
  abundance_rel <- abundance_long %>%
    dplyr::left_join(sample_total_tpm_all, by = "sample_short") %>%
    mutate(relative_abundance = TPM / total_tpm_all)
  
  rare_votus_list <- abundance_rel %>%
    filter(relative_abundance < 0.001 & relative_abundance > 0) %>%
    pull(vOTU_id) %>%
    unique()
}

message(sprintf("✅ Rare vOTUs: %d\n", length(rare_votus_list)))

# ============================================================================
# STEP 4: Filter host predictions for rare vOTUs
# ============================================================================
message("🔗 Filtering host predictions for rare vOTUs...")

host_rare <- host_filtered %>%
  filter(vOTU_id %in% rare_votus_list)

message(sprintf("✅ Rare vOTUs with host predictions: %d\n", n_distinct(host_rare$vOTU_id)))

# ============================================================================
# STEP 5: Prepare abundance data for rare vOTUs
# ============================================================================
message("📊 Preparing abundance data...")

# Sample mapping (exactly as old code)
sample_map <- data.frame(
  sample_id = c("FCH3CGLCCXYL2DAA", "FCH3CGLCCXYL2EAA", "FCH3CGLCCXYL4FAA", "TARATASF"),
  sample_source1 = c("SA", "IA", "BS", "DA"),
  stringsAsFactors = FALSE
)

# Extract TPM for rare vOTUs
tpm_mat <- assays(tse)$tpm

tpm_rare <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  filter(vOTU_id %in% rare_votus_list) %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "tpm") %>%
  dplyr::left_join(sample_map, by = "sample_id") %>%
  filter(!is.na(sample_source1))

message("✅ Abundance data prepared\n")

# ============================================================================
# STEP 6: Create heatmap data (presence/absence by Phylum)
# ============================================================================
message("🗂️ Creating heatmap data...")

# Combine host predictions with abundance (many-to-many expected)
host_abundance <- host_rare %>%
  dplyr::left_join(tpm_rare, by = "vOTU_id", relationship = "many-to-many")

# Calculate presence/absence per sample and Phylum
heatmap_data <- host_abundance %>%
  filter(tpm > 0) %>%
  group_by(sample_source1, Phylum) %>%
  summarise(
    n_vOTUs = n_distinct(vOTU_id),
    presence = 1,
    .groups = "drop"
  )

# Complete the grid (all samples × all phyla)
all_samples <- c("BS", "SA", "IA", "DA")
all_phyla <- unique(host_filtered$Phylum)

heatmap_data_complete <- expand.grid(
  sample_source1 = all_samples,
  Phylum = all_phyla,
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(heatmap_data, by = c("sample_source1", "Phylum")) %>%
  mutate(
    n_vOTUs = if_else(is.na(n_vOTUs), 0, as.numeric(n_vOTUs)),
    presence = if_else(is.na(presence), 0, presence)
  )

message(sprintf("✅ Heatmap data ready: %d phyla detected\n", sum(heatmap_data_complete$presence > 0)))

# ============================================================================
# STEP 7: Create heatmap plot (exactly matching old code)
# ============================================================================
message("🎨 Creating heatmap...")

# Set factor levels (exactly as old code)
heatmap_data_complete$sample_source1 <- factor(
  heatmap_data_complete$sample_source1, 
  levels = c("BS", "SA", "IA", "DA")
)

# Order phyla by total presence (exactly as old code)
phylum_order <- heatmap_data_complete %>%
  group_by(Phylum) %>%
  summarise(total_presence = sum(presence), .groups = "drop") %>%
  arrange(desc(total_presence)) %>%
  pull(Phylum)

heatmap_data_complete$Phylum <- factor(heatmap_data_complete$Phylum, levels = phylum_order)

# Create plot (exactly matching old code style)
p_heatmap <- ggplot(heatmap_data_complete, aes(x = sample_source1, y = Phylum)) +
  geom_tile(aes(fill = factor(presence)), color = "white", linewidth = 1) +
  scale_fill_manual(
    values = c("0" = "gray90", "1" = "#66C2A5"),
    guide = "none"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 18, family = "Times", color = "black"),
    axis.text.y = element_text(color = "black", size = 18, family = "Times", face = "italic"),
    axis.title = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p_heatmap)

# Save plot
ggsave(path_target("FigS5c_host_heatmap.pdf"), plot = p_heatmap, width = 8, height = 10, dpi = 300)
ggsave(path_target("FigS5c_host_heatmap.png"), plot = p_heatmap, width = 8, height = 10, dpi = 300)

# Save data
write_csv(heatmap_data_complete, path_target("FigS5c_host_heatmap_data.csv"))

# Summary statistics
summary_data <- heatmap_data_complete %>%
  group_by(sample_source1) %>%
  summarise(
    n_phyla_present = sum(presence > 0),
    n_vOTUs_total = sum(n_vOTUs),
    .groups = "drop"
  )

write_csv(summary_data, path_target("FigS5c_summary.csv"))

message("✅ Figure S5c completed and saved!\n")

# Print summary
cat("\n========================================================================\n")
cat("📊 Summary Statistics:\n")
cat("========================================================================\n")
print(summary_data)
cat("========================================================================\n\n")

```


### Task 4: Figure S5b (OLD CODE VERSION - GLOBAL Definition)
```{r}
#| label: figs5c-host-heatmap-OLD-GLOBAL
#| message: true
#| fig.width: 8
#| fig.height: 10

message("========================================================================")
message("Figure S5c: OLD CODE VERSION - GLOBAL Definition")
message("Confidence threshold: ≥75")
message("Criterion: Relative Abundance < 0.1% GLOBALLY")
message("========================================================================\n")

# ============================================================================
# STEP 1: Load host prediction data from TSE metadata
# ============================================================================
message("📥 Loading host prediction data from TSE metadata...")

host_genome <- metadata(tse)$host_genome_edges

stopifnot(!is.null(host_genome))

message("✅ host_genome loaded: ", nrow(host_genome), " predictions")
message("   Columns: ", paste(colnames(host_genome), collapse = ", "), "\n")

# ============================================================================
# STEP 2: Filter high-confidence host predictions (≥75)
# ============================================================================
CONFIDENCE_THRESHOLD <- 80

message(sprintf("🔍 Filtering host predictions (confidence ≥ %d)...\n", CONFIDENCE_THRESHOLD))

host_filtered <- host_genome %>%
  filter(Confidence.score >= CONFIDENCE_THRESHOLD) %>%
  mutate(
    Phylum = str_extract(Host.taxonomy, "(?<=p__)[^;]+"),
    Phylum = if_else(is.na(Phylum), "Unclassified", Phylum)
  )

message(sprintf("✅ High-confidence predictions: %d rows", nrow(host_filtered)))
message(sprintf("✅ Unique phyla: %d\n", n_distinct(host_filtered$Phylum)))

# ============================================================================
# STEP 3: Calculate GLOBAL rare vOTUs (< 0.1% globally)
# ============================================================================
message("📊 Calculating GLOBAL relative abundance...")

# Sample mapping (exactly as old code)
sample_map <- data.frame(
  sample_id = c("FCH3CGLCCXYL2DAA", "FCH3CGLCCXYL2EAA", "FCH3CGLCCXYL4FAA", "TARATASF"),
  sample_source1 = c("SA", "IA", "BS", "DA"),
  stringsAsFactors = FALSE
)

# Extract TPM
tpm_mat <- assays(tse)$tpm

# Convert to long format
abundance_long <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "TPM") %>%
  dplyr::left_join(sample_map, by = "sample_id")

# *** GLOBAL DEFINITION: Calculate TOTAL TPM across ALL samples ***
total_TPM_global <- sum(abundance_long$TPM, na.rm = TRUE)

message(sprintf("✅ Total TPM (GLOBAL): %.2f\n", total_TPM_global))

# Calculate abundance per vOTU per sample_source1
abundance_per_group <- abundance_long %>%
  group_by(sample_source1, vOTU_id) %>%
  summarise(TPM_sum = sum(TPM, na.rm = TRUE), .groups = "drop")

# *** Calculate GLOBAL relative abundance ***
abundance_per_group <- abundance_per_group %>%
  mutate(rel_abundance_global = TPM_sum / total_TPM_global)

# Identify RARE vOTUs (< 0.1% GLOBALLY and > 0)
rare_votus_list_global <- abundance_per_group %>%
  filter(rel_abundance_global < 0.001 & rel_abundance_global > 0) %>%
  pull(vOTU_id) %>%
  unique()

message(sprintf("✅ Identified %d rare vOTUs (< 0.1%% GLOBALLY)\n", length(rare_votus_list_global)))

# ============================================================================
# STEP 4: Filter host predictions for rare vOTUs
# ============================================================================
message("🔗 Filtering host predictions for rare vOTUs...")

host_rare <- host_filtered %>%
  filter(vOTU_id %in% rare_votus_list_global)

message(sprintf("✅ Rare vOTUs with host predictions: %d\n", n_distinct(host_rare$vOTU_id)))

# ============================================================================
# STEP 5: Prepare abundance data for rare vOTUs
# ============================================================================
message("📊 Preparing abundance data...")

# Extract TPM for rare vOTUs
tpm_rare <- as.data.frame(tpm_mat) %>%
  rownames_to_column("vOTU_id") %>%
  filter(vOTU_id %in% rare_votus_list_global) %>%
  pivot_longer(cols = -vOTU_id, names_to = "sample_id", values_to = "tpm") %>%
  dplyr::left_join(sample_map, by = "sample_id") %>%
  filter(!is.na(sample_source1))

message("✅ Abundance data prepared\n")

# ============================================================================
# STEP 6: Create heatmap data (presence/absence by Phylum)
# ============================================================================
message("🗂️ Creating heatmap data...")

# Combine host predictions with abundance (many-to-many expected)
host_abundance <- host_rare %>%
  dplyr::left_join(tpm_rare, by = "vOTU_id", relationship = "many-to-many")

# Calculate presence/absence per sample and Phylum
heatmap_data <- host_abundance %>%
  filter(tpm > 0) %>%
  group_by(sample_source1, Phylum) %>%
  summarise(
    n_vOTUs = n_distinct(vOTU_id),
    presence = 1,
    .groups = "drop"
  )

# Complete the grid (all samples × all phyla)
all_samples <- c("BS", "SA", "IA", "DA")
all_phyla <- unique(host_filtered$Phylum)

heatmap_data_complete <- expand.grid(
  sample_source1 = all_samples,
  Phylum = all_phyla,
  stringsAsFactors = FALSE
) %>%
  dplyr::left_join(heatmap_data, by = c("sample_source1", "Phylum")) %>%
  mutate(
    n_vOTUs = if_else(is.na(n_vOTUs), 0, as.numeric(n_vOTUs)),
    presence = if_else(is.na(presence), 0, presence)
  )

message(sprintf("✅ Heatmap data ready: %d phyla detected\n", sum(heatmap_data_complete$presence > 0)))

# ============================================================================
# STEP 7: Create heatmap plot
# ============================================================================
message("🎨 Creating heatmap...")

# Set factor levels
heatmap_data_complete$sample_source1 <- factor(
  heatmap_data_complete$sample_source1, 
  levels = c("BS", "SA", "IA", "DA")
)

# Order phyla by total presence
phylum_order <- heatmap_data_complete %>%
  group_by(Phylum) %>%
  summarise(total_presence = sum(presence), .groups = "drop") %>%
  arrange(desc(total_presence)) %>%
  pull(Phylum)

heatmap_data_complete$Phylum <- factor(heatmap_data_complete$Phylum, levels = phylum_order)

# Create plot
p_heatmap_old <- ggplot(heatmap_data_complete, aes(x = sample_source1, y = Phylum)) +
  geom_tile(aes(fill = factor(presence)), color = "white", linewidth = 1) +
  scale_fill_manual(
    values = c("0" = "gray90", "1" = "#66C2A5"),
    guide = "none"
  ) +
  labs(title = "OLD CODE VERSION - GLOBAL Definition") +
  theme_minimal(base_size = 14) +
  theme(
    text = element_text(family = "Times"),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 18, family = "Times", color = "black"),
    axis.text.y = element_text(color = "black", size = 18, family = "Times", face = "italic"),
    axis.title = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5, color = "red"),
    legend.position = "none",
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

print(p_heatmap_old)

# Save plot
ggsave(path_target("FigS5c_host_heatmap_OLD_GLOBAL.pdf"), plot = p_heatmap_old, width = 8, height = 10, dpi = 300)
ggsave(path_target("FigS5c_host_heatmap_OLD_GLOBAL.png"), plot = p_heatmap_old, width = 8, height = 10, dpi = 300)

# Save data
write_csv(heatmap_data_complete, path_target("FigS5c_host_heatmap_data_OLD_GLOBAL.csv"))

# Summary statistics
summary_data <- heatmap_data_complete %>%
  group_by(sample_source1) %>%
  summarise(
    n_phyla_present = sum(presence > 0),
    n_vOTUs_total = sum(n_vOTUs),
    .groups = "drop"
  )

write_csv(summary_data, path_target("FigS5c_summary_OLD_GLOBAL.csv"))

# Save rare vOTU list for comparison
write_csv(
  data.frame(vOTU_id = rare_votus_list_global, definition = "GLOBAL"),
  path_target("rare_vOTUs_S5c_OLD_GLOBAL.csv")
)

message("✅ Figure S5c OLD CODE VERSION completed and saved!\n")

# Print summary
cat("\n========================================================================\n")
cat("📊 Summary Statistics (GLOBAL Definition):\n")
cat("========================================================================\n")
cat("Total TPM (GLOBAL):", sprintf("%.2f", total_TPM_global), "\n")
cat("Rare vOTUs (< 0.1% GLOBALLY):", length(rare_votus_list_global), "\n")
cat("\n")
print(summary_data)
cat("========================================================================\n\n")


```

```{r}

```

