---
title: "11-figS6-amg-DA"
title-block-banner: true
author:
  - name: Jinlong Ru
    orcid: 0000-0002-6757-6018
date: 2025-12-20
toc: true
toc-depth: 4
number-sections: true
code-fold: true
code-line-numbers: true
code-tools: true
format: 
  html:
    embed-resources: true
    smooth-scroll: true
    page-layout: full
reference-location: section
citation-location: document
params:
  name: "11-figS6-amg-DA"
---

**Updated: `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'CET')` CET.**

The purpose of this document is ...

```{r}
#| label: params
#| eval: !expr interactive()
#| include: false
params = list(name = "11-figS6-amg-DA")
```

```{r}
#| label: setup
#| message: false
#| include: false
#| warning: false
wd <- "analyses"
if (basename(getwd()) != wd) {
  setwd(here::here(wd))
}
here::i_am(paste0(params$name, ".qmd"), uuid = "ee61afbe-a077-4910-9d6a-35bacc1115ab")
projthis::proj_create_dir_target(params$name, clean = FALSE)
path_target <- projthis::proj_path_target(params$name)
path_source <- projthis::proj_path_source(params$name)
path_raw <- path_source("00-raw")
path_resource <- here::here(path_raw, "d00-resource")
path_data <- here::here(path_raw, paste0("d", params$name))
dir.create(path_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(path_data, recursive = TRUE, showWarnings = FALSE)
dir.create(path_resource, recursive = TRUE, showWarnings = FALSE)
```

```{r}
suppressPackageStartupMessages({
  library(here)
  library(tidyverse)
  library(pheatmap)
  library(RColorBrewer)
  library(openxlsx)
})
```

## Tasks

This section is designated for your workflow implementation. **Please write all your analysis and processing code here.**

Maintaining your code within this section ensures a clean project structure and facilitates reproducibility. We recommend organizing your workflow into modular tasks—such as data cleaning, analysis, and visualization—using distinct code chunks.

Below is an example structure to get you started:



### Task 1: Plot Top 20 AMG Heatmap (Ranked by DA)

```{r}
#| label: plot-amg-heatmap-da
#| fig.width: 8
#| fig.height: 10

# Read deduplicated pathway table from 05-fig5-heatmap
dedup_path <- path_source("05-fig5-heatmap", "deduplicated_pathway_table.xlsx")

if (!file.exists(dedup_path)) {
  stop("deduplicated_pathway_table.xlsx not found in 05-fig5-heatmap outputs")
}

df <- openxlsx::read.xlsx(dedup_path)

# Clean data (key column names: dbid, vOTU_id, sample_group, TPM, db_desc)
df_clean <- df %>%
  dplyr::select(dbid, vOTU_id, sample_group, TPM, db_desc) %>%
  dplyr::filter(!is.na(dbid), !is.na(sample_group), !is.na(TPM)) %>%
  dplyr::distinct()

message("Clean data: ", nrow(df_clean), " rows")

# Special gene name mapping (same as Fig5b)
special_map <- tibble::tribble(
  ~dbid, ~gene_symbol,
  "K00525", "nrdB",
  "EC:1.17.4.1", "nrdB",
  "K00973", "rfbA",
  "K01710", "rfbB",
  "K00560", "thyA",
  "K01520", "dut",
  "K00558", "DNMT1",
  "K17398", "DNMT3A",
  "K00287", "folA"
)

# Extract gene labels
df_labeled <- df_clean %>%
  dplyr::left_join(special_map, by = "dbid") %>%
  dplyr::mutate(
    auto_label = str_remove_all(db_desc, "\\[.*?\\]") %>%
      str_split(";", simplify = TRUE) %>%
      .[, 1] %>%
      str_trim() %>%
      str_split(",", simplify = TRUE) %>%
      .[, 1] %>%
      str_trim(),
    gene_symbol = if_else(is.na(gene_symbol), auto_label, gene_symbol)
  ) %>%
  dplyr::select(-auto_label)

# Summarize TPM per gene and sample
gene_summary <- df_labeled %>%
  dplyr::group_by(gene_symbol, sample_group) %>%
  dplyr::summarise(Total_TPM = sum(TPM, na.rm = TRUE), .groups = "drop")

# Convert to wide format
gene_matrix <- gene_summary %>%
  tidyr::pivot_wider(names_from = sample_group, values_from = Total_TPM, values_fill = 0)

# Select Top 20 by DA abundance (key difference!)
samples <- c("BS", "SA", "IA", "DA")
available_samples <- intersect(samples, colnames(gene_matrix))

gene_top20_da <- gene_matrix %>%
  dplyr::arrange(desc(DA)) %>%
  dplyr::slice_head(n = 20)

message("Top 20 AMG genes (ranked by DA TPM)")

# log10 transformation
gene_log_matrix_da <- gene_top20_da %>%
  dplyr::select(gene_symbol, all_of(available_samples)) %>%
  tibble::column_to_rownames("gene_symbol") %>%
  dplyr::mutate(across(everything(), ~log10(.x + 1)))

gene_log_matrix_da <- gene_log_matrix_da[, available_samples]

# Set row names to italic format
rownames(gene_log_matrix_da) <- paste0("italic('", rownames(gene_log_matrix_da), "')")

# Color scheme
my_colors <- colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))(100)

# Plot pheatmap
p_amg_da <- pheatmap::pheatmap(
  gene_log_matrix_da,
  color = my_colors,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_colnames = TRUE,
  show_rownames = TRUE,
  labels_row = parse(text = rownames(gene_log_matrix_da)),
  main = "",
  fontsize_row = 10,
  fontsize_col = 18,
  fontsize = 13,
  fontfamily = "Times",
  angle_col = 0,
  cellwidth = 35,
  cellheight = 12,
  border_color = "gray80"
)

# Save figures
ggsave(path_target("FigS6_top20_AMG_DA_heatmap.png"), 
       plot = p_amg_da, width = 8, height = 10, dpi = 300)
ggsave(path_target("FigS6_top20_AMG_DA_heatmap.pdf"), 
       plot = p_amg_da, width = 8, height = 10)

# Save data
write.csv(
  gene_log_matrix_da %>% tibble::rownames_to_column("Gene"),
  path_target("top20_AMG_DA_log10TPM_matrix.csv"), 
  row.names = FALSE
)

openxlsx::write.xlsx(
  gene_top20_da, 
  path_target("top20_AMG_DA_raw.xlsx"), 
  rowNames = FALSE
)

message("✅ FigS6 completed: ", nrow(gene_log_matrix_da), " genes")

```



